An Anatomical and Physiological Basis for Flexible Coincidence Detection in the Auditory System

Version 1

Abstract

Animals navigate the auditory world by recognizing complex sounds, from the rustle of a predator to the call of a potential mate. This ability depends in part on the octopus cells of the auditory brainstem, which respond to multiple frequencies that change over time, as occurs in natural stimuli. Unlike the average neuron, which integrates inputs over time on the order of tens of milliseconds, octopus cells must detect momentary coincidence of excitatory inputs from the cochlea during an ongoing sound on both the millisecond and submillisecond time scale. Here, we show that octopus cells receive inhibitory inputs on their dendrites that enhance opportunities for coincidence detection in the cell body, thereby allowing for responses both to rapid onsets at the beginning of a sound and to frequency modulations during the sound. This mechanism is crucial for the fundamental process of integrating the synchronized frequencies of natural auditory signals over time.

Introduction

Perception depends on the ability of neurons to encode discrete features of complex external stimuli. These computations are determined by the type of information received from the periphery, the nature and position of the synapses, and the biophysical properties of the target neuron. For instance, in the auditory system, spiral ganglion neurons (SGNs) encode sound information from hair cells in the cochlea and distribute it to specialized target neurons in the cochlear nucleus complex (CNC) that extract individual elements of the original sound, such as frequency, phase, intensity, and timing. The outcome of each dedicated computation continues to be processed through parallel ascending pathways and ultimately to the cortex, where the auditory features are reassembled to generate a percept. Thus, understanding how these first computations are made is a key step towards deciphering the basis of perception.

To generate accurate percepts of complex auditory stimuli, neurons must compute both what frequencies are present and when those frequencies occur across multiple time scales. For instance, overlapping sound stimuli, such as two competing speakers in a noisy room, are perceptually distinguished by correctly binding together frequencies with coherent onsets that then continue to change together over time1,2. Such computations require coincidence detection that can accurately encode co-occurring frequencies with submillisecond precision. Frequency information is communicated by SGNs, whose auditory nerve fibers (ANFs) project through the eighth nerve, bifurcate, and spread tonotopically to fill each division of the CNC (Fig. 1A). In addition, SGNs fall into physiologically distinct subtypes that are recruited at different intensities, thereby allowing sounds to be detected across a wide dynamic range and in the presence of background noise. Many target neurons receive SGN inputs from a limited range of frequencies and fire at the onset of the sound, effectively breaking it up into its frequency components. This presents a challenge for perception as auditory circuits must ultimately bind together co-occurring frequencies while also retaining information about their sequence to locate and recognize sounds.

Excitatory and inhibitory synapses to octopus cells form two domains.

(A) Illustration of spiral ganglion neuron (SGN) central axons branching within a parasagittal section of the mouse cochlear nucleus complex (CNC). SGN somas in the cochlea are tono-topically organized according to frequency. Axons remain organized throughout the ventral (VCN) and dorsal (DCN) divisions of the CNC. Octopus cells (inset) are found in the octopus cell area (OCA) of the VCN. (B) Excitatory SGN puncta labeled with a VGLUT1 antibody (left) and glycinergic puncta labeled with Glyt2Cre-dependent syp/tdT (Glyt2-syp/tdT; right) in a par-asagittal section of the CNC. The teardrop-shaped OCA is not devoid of inhibitory inputs, although less prominent than in the surrounding CNC. (C) A Thy1 sparsely labeled octopus cell with excitatory SGN (VGLUT1) and inhibitory (Glyt2-syp/tdT) puncta. Micrographs of 3µm confocal z-stacks show puncta on the medial surface of a soma (top) and a dendrite (bottom). (D) Representative reconstruction of excitatory SGN puncta labeled with Foxg1Cre-dependent syp/tdT (Foxg1-syp/tdT; blue) onto a Thy1+ octopus cell. (E) Representative reconstruction of inhibitory puncta labeled with Glyt2-syp/tdT (green) onto a Thy1 sparsely labeled octopus cell (white). (F) Puncta density for excitatory SGN (Foxg1-syp/tdT; black: 10.7 ± 3.0, n = 8 cells, 4 mice) and glycinergic puncta (Glyt2-syp/tdT; green: 4.2 ± 0.8, n = 8 cells, 3 mice) on octopus cell dendrites. Data are presented as mean ± SD. (G) Puncta density on somas for excitatory SGN (black: 13.3 ± 2.2, n = 8 cells, 4 mice) and glycinergic puncta (green: 1.8 ± 0.1, n = 8 cells, 3 mice) and the density along the length of dendrites. Data are presented as mean ± SEM. (H) Top: Illustration of an octopus cell and the ratio between excitatory SGN puncta and glycinergic puncta. Inset: Illustration of an octopus cell and the relative innervation densities of excitatory SGNs (blue) and inhibitory puncta (green).

In the mammalian auditory system, precise encoding of broadband timing information begins with the octopus cells of the CNC. Octopus cells are excitatory neurons that bind together co-occurring frequency information on a submillisecond timescale and send this information along one of the parallel ascending pathways in the auditory brainstem. Octopus cells are named for their large-diameter tentacle-like dendrites3,4, which are oriented unidirectionally across a tonotopic array of SGNs such that each neuron integrates inputs from a wide range of frequencies5–8. SGNs provide the major excitatory inputs onto octopus cells. Biophysically, octopus cells have low input resistances near rest (∼4MΩ), fast time constants (∼200µs), and large low-voltage-activated potassium (∼40nS at rest) and hyperpolarization activated (∼62nS at rest) conductances. Together these properties give these cells impressively narrow windows of coincidence detection on the order of 1 millisecond9–16. This combination of receiving SGN innervation across broad frequencies and their biophysical specializations establish octopus cells as spectrotemporal coincidence detectors that can reliably encode the timing of complex stimuli, such as the broadband transients found in speech and other natural sounds13,17,18. Fittingly, in vivo recordings from octopus cells demonstrate their ability to phase lock to broadband transients at rates up to 1kHz19–21. Moreover, computational models of octopus cells demonstrate that onset responses are governed by the cell’s biophysical specializations and are, in large part, the result of temporal summation of excitation22–28. The simplicity of its connectivity combined with the precision of its temporal computations makes the octopus cell an attractive model for understanding how specialized anatomical and electrophysiological properties contribute to neuronal computations.

Although the octopus cell’s integration of SGN inputs within a very narrow time frame enables canonical coincidence detection, such a model does not explain how other temporal features of sound stimuli are encoded. Indeed, octopus cells encode spectrotemporal sequences within their broadly-tuned response areas, like frequency modulated sounds, that likely require further circuit specializations29. Although threshold somatic depolarization can be sufficient to activate an octopus cell23, the vast majority of synapses are found on dendrites. Further, SGN inputs are organized tonotopically along octopus cell dendrites, with inputs from high frequency regions located more distal than those from low frequency regions. Dendritic morphology, passive cable properties, active resting membrane properties, and the spatial and temporal relationship between synaptic inputs can all impact excitatory post synaptic potential (EPSP) summation as excitation sweeps across the dendritic arbor and towards the soma. This raises the possibility that computations made in the dendrites influence the effective window of coincidence detection by the octopus cell. Such mechanisms could enable flexible processing that is adaptive to the dynamics of the environment without compromising the fidelity of high-fidelity coincidence computations.

Here, we sought to define the circuit mechanisms that allow octopus cells to act as coincidence detectors across time scales. We generated a comprehensive anatomical and physiological map of excitatory and synaptic inputs onto octopus cell somas and dendrites and examined how this circuit organization influences octopus cell activation. Through a combination of in vitro experiments and computational modeling, we show that the somatic summation of excitation is shaped by dendritic inhibition. Thus, octopus cells depend on compartmentalized computations that enable preservation of timing information both at the moment of stimulus onset and within an extended window for evidence accumulation, which is fundamental for the spectrotemporal integration of natural auditory stimuli.

Results

The balance of excitatory and inhibitory synapses is different in somatic and dendritic domains

To determine the wiring pattern that drives octopus cell computations, we generated a detailed map of excitatory and inhibitory synaptic inputs (Fig. 1). Overall, octopus cells receive abundant excitatory VGLUT1+ innervation from SGNs30,31 and sparse inhibitory innervation from glycinergic neurons, as visualized using the glycinergic Cre driver Glyt2Cre and the Ai34 synaptophysin-tdTomato (syp/tdT) fusion protein reporter (Fig. 1B). In addition, sparse inhibitory inputs nestle between SGN inputs, especially on octopus cell dendrites (Fig. 1C).

Quantification of the number and distribution of presynaptic puncta onto octopus cells revealed marked differences in the ratio of excitation and inhibition in the somatic and dendritic compartments. Since innervation patterns have never been systematically analyzed, we made three-dimensional reconstructions of 16 octopus cells and their excitatory SGN (n = 8 cells, 4 mice) and inhibitory (n = 8 cells, 3 mice) inputs. Octopus cells were visualized using a Thy1 reporter and presynaptic puncta were labeled with the syp/tdT reporter driven either by Foxg1Cre (Fig. 1D) or Glyt2Cre (Fig. 1E). Consistent with qualitative assessment, the density of SGN inputs was higher (10.7 ± 3.0 SGN puncta/100µm2) than that of inhibitory inputs (4.2 ± 0.8 puncta/100µm2, Fig. 1F). Moreover, the relative proportions of excitatory and inhibitory inputs differed in the soma and dendrites (Fig. 1G). On somas, SGNs provided dense innervation that continued on the proximal dendrite, then gradually declined with distance from the soma. By contrast, somas received very few inhibitory inputs. On dendrites, inhibitory puncta were evenly distributed. As a result, octopus cells have a strikingly different average ratio of excitatory and inhibitory puncta on the soma (7:1) and on the dendrite (5:2), suggesting that each compartment contributes differentially to the final computation made by the octopus cell (Fig. 1H).

The majority of excitatory synapses on octopus cells are from type Ia SGNs

Although uniformly glutamatergic, SGNs exhibit stereotyped physiological differences in response thresholds that could affect the nature of their inputs onto octopus cells and influence octopus cell coincidence detection32–35,35,36. It was recently discovered that there are three molecularly distinct SGN subtypes, referred to as Ia, Ib, and Ic SGNs, which correlate with previously shown physiological groups37–43 (Fig. 2A). Therefore, we further categorized excitatory inputs based on SGN subtype identity. These can be identified with the presence of Ntng1-dependent reporter expression44 in Ib and Ic SGNs (Ib/c) and its absence in Ia SGNs40,41,43, coupled with very low to undetectable levels of calretinin (CR-) in Ic SGNs, and moderate to high levels of calretinin (CR+, CR++) in Ib and Ia SGNs (Fig. 2B). Ntng1Cre-labeled Ib/c SGNs accounted for 60.1 ± 2.6% of the entire population, with 28.5 ± 12.2% Ib SGNs, 31.6% Ic SGNs, and 39.9 ± 2.6% Ia SGNs (Fig. 2C: n = 1599 neurons, 4 mice; mean ± SD). Further, proportions of SGN subtypes matched scRNA-seq estimates (Fig. 2C), indicating that this approach provides full coverage. SGN subtype identity was further confirmed by examining the spatial organization of SGN peripheral processes en route to the IHCs in the cochlea (Fig 2A, Supp. Fig. 1A-C).

Type Ia SGNs are the primary excitatory contributors to octopus cells.

(A) Ia (yellow), Ib (orange), and Ic (magenta) SGN axons innervate the CNC. SGNs have a continuum of properties organized along the pillar-modiolar axis of inner hair cells (IHCs) and the habenula. Ntng1-expressing Ib/c fibers are positioned on the modiolar side (closest to the ganglion). Strongly calretinin immunopositive Ia fibers are on the other side (closest to the pillar cells). This organization correlates with spontaneous rates (SR) and thresholds (thresh.) measured in vivo. Somas of all SGN subtypes are found at all tonotopic locations. (B) Calretinin (CR) immunolabeling distinguishes SGN subtypes. Ia/b somas label with high (CR++) and medium (CR+) levels of CR, respectively. Ic somas label with low to undetectable levels of CR (CR-). Ntng1Cre-mediated expression of tdT (Ntng1-tdT) labels Ib/c SGNs. (C) Ias (tdT-, CR++) make up 39.9 ± 2.6% of the SGN population. Ib/cs (tdT+) make up 60.1 ± 2.6% of the SGN population. Ibs (tdT+CR+) make up 28.5 ± 12.2% of the SGN population (n = 1599 neurons, 4 mice). Data are mean ± SD; individual points represent percent coverage per animal, lines connect measurements from the same animal. Dotted lines indicate percentages from 1: Petitpré et al. 2018, 2: Shrestha et al. 2018, and 3: Sun et al. 2018. (D) Reconstructed Ib/c puncta labeled with Ntng1Cre-dependent syp/tdT (Ntng1-syp/tdT; magenta) onto a Thy1+ octopus cell. (E) Puncta density for all SGNs (Foxg1-syp/tdT; black: data from Fig. 1F), Ias (Foxg1 -Ntng1; yellow: 6.6 ± 1.0), and Ib/cs (Ntng1-syp/tdT; magenta: 4.1 ± 1.0, n = 9 cells, 5 mice) along the total dendritic length. Ia density was calculated by subtracting Ib/c density from total SGN density; lines connect measurements from the same reconstruction. Data are mean ± SD. (F) Puncta density on somas for all SGNs (Foxg1-syp/tdT; black: data from Fig. 1E), Ias (Foxg1 -Ntng1; yellow: 8.4 ± 2.3), and Ib/cs (Ntng1-syp/tdT; magenta: 4.9 ± 1.2, n = 9 cells, 5 mice) and the density along the length of dendrites. Data are mean ± SEM. (G) Relative innervation densities of Ias (yellow), Ib/cs (magenta), and inhibitory puncta (green).

Within the ventral cochlear nucleus (VCN), where octopus cells reside, Ntng1Cre labeling was restricted to SGNs (Supp. Fig. 1D). In the VCN, expression of Ntng1-tdT and CR in SGN central axons was consistent with the moderate to undetectable levels of CR in Ntng1-tdT SGN somas in the periphery (Supp. Fig. 2E). Thus, Ntng1Cre-driven expression of syp/tdT is an appropriate tool for mapping subtype-specific connectivity onto octopus cells.

Reconstruction of Ntng1Cre-labeled Ib/c puncta (Fig. 2D) demonstrated that octopus cells are dominated by inputs from Ia SGN fibers, which are the fibers with the lowest response thresholds and highest rates of spontaneous activity. Octopus cell dendrites received 4.1 ± 1.0 puncta/100µm2 from Ib/c SGNs (Fig. 2E, magenta: n = 9 cells, 5 mice; mean ± SD), accounting for 38% of the total SGN density. Given that Ntng1-tdT+ cells account for 60.1% of the SGN population (Fig. 2C, magenta), Ib/c inputs were underrepresented on octopus cells. Octopus cells receive similarly low innervation from Ic inputs (Supp. Fig. 2G-I: n = 6 cells, 2 mice), as estimated from the degree of sparse labeling of Ic axons achieved by Myo15iCre reporter expression (Supp. Fig. 2A-F) and the expected proportion of Ic SGNs in the ganglion (Supp. Fig. 2F). By contrast, Ia SGNs, which comprise only ∼40% of the total population (Fig. 2C, yellow), accounted for 62 ± 9.7% of SGN synapses on octopus cells (Fig. 2E, yellow: 6.6 ± 1.0 puncta/100µm2). All three subtypes showed the same overall distribution from the soma to the distal dendrite (Fig. 2F). Together, excitatory and inhibitory puncta densities in the innervation maps indicate the average octopus cell receives ∼1035 SGN synapses (642 Ia SGN, 393 Ib/c SGN) and ∼354 inhibitory synapses. Additionally, the majority of synapses on the average octopus cell (83%) are found on dendrites, highlighting their critical role in the octopus cell computation.

Octopus cell reconstructions showed the same basic wiring patterns regardless of where each cell was positioned in the octopus cell area. The tonotopic position of all reconstructed octopus cell somas was estimated in 3D reconstructions aligned to a normalized CNC model of tonotopy45. Octopus cells had similar morphologies (Supp. Fig. 3E-G) and patterns of synaptic innervation (Supp. Fig. 3H-M) regardless of where they were positioned along the tonotopic axis. Thus, we have established a wiring diagram that finds low threshold, Ia SGN synapses to be the primary input to both the soma and dendrites of octopus cells. Additionally, the whole-neuron wiring diagram identifies a dendritic domain where inhibitory synapses are approximately equal in number to the Ib/c excitatory synapses from the periphery.

SGN subtype inputs to octopus cells do not differ in short term plasticity.

(A) Illustration of the experimental paradigm and representative EPSPs recorded during in vitro whole-cell current clamp recordings of octopus cells. SGN stimulation method included electrical stimulation or full-field, light-evoked activation of Foxg-ChR2 or Ntng1-ChR2 SGNs. TTL trigger pulses are shown in gray. (B) Paired pulse ratios for electrically stimulated SGNs (open circles: n = 5 cells, 3 mice), ChR2 stimulated SGNs (Foxg1-ChR2; black: n = 8 cells, 5 mice), and ChR2 stimulated Ib/c SGNs (Ntng1-ChR2; magenta: n = 7 cells, 6 mice) at three interstimulus intervals. With electrical stimulation, SGN inputs to octopus cells were stable and exhibited slight facilitation at 50 Hz (20ms interstimulus interval). ChR2 stimulation caused paired pulse depression not seen with electrical stimulation. Data are presented as mean ± SD. Each data point represents the average paired pulse ratio for a cell. p values from ANOVA and subsequent Tukey HSD test are reported for comparisons between methods of SGN activation (electrical and ChR2) and SGN subpopulation composition within method of activation (SGN-ChR2 and Ib/c-ChR2). Welch’s ANOVA was used for comparisons at 20ms interstimulus interval (50Hz) as data in this condition did not meet the homogeneity of variance assumption. (C) Pulse ratios at 50Hz for electrically stimulated SGNs with physiological 1.4mM Ca2+ ACSF (open circles: n = 5 cells, 3 mice), electrically stimulated SGNs with 2.4mM Ca2+ ACSF (grey: n = 3 cells, 2 mice) and ChR2 stimulated SGNs with physiological ACSF (Foxg1-ChR2; black: n = 8 cells, 5 mice). Data are presented as mean ± SD. p < 0.001 from ANOVA and subsequent Tukey HSD test for all comparisons between methods of SGN activation (electrical and ChR2). There were no statistically significant differences for all comparisons under 1.4mM and 2.4mM Ca2+ (p > 0.100, ANOVA).

SGN inputs to octopus cells facilitate at high stimulation frequencies

Whether or not an octopus cell responds to its inputs depends on when and how EPSPs travel to and then summate in the soma. To determine if SGN subtypes transmit information differently to their central targets, we performed in vitro whole-cell current clamp recordings of octopus cells (Fig. 3A) while using Channelrhodopsin-2 (ChR2) to stimulate either all SGNs (Foxg1-ChR2) or only Ib/c SGNs (Ntng1-ChR2). Trains of ChR2-evoked SGN stimulation, ranging from 5 to 50Hz, in both the total SGN population (Fig. 3B, black: n = 8 cells, 5 mice) and the Ib/c SGN population (Fig. 3B, magenta: n = 7 cells, 6 mice) exhibited no differences in paired-pulse plasticity at any frequency of stimulation (p > 0.35 at all interstimulus intervals, Tukey’s HSD), although the Ib/c population exhibited higher variability than the total SGN population (at 20ms: SD = 0.11, SD = 0.24, respectively).

ChR2-evoked synaptic responses are known to artificially undergo synaptic depression46,47. To determine if the paired pulse depression measured in ChR2-stimulated experiments was physiological (Fig. 3B), we used electrical stimulation to evoke EPSPs from SGNs. Electrically-evoked SGN EPSPs had higher paired-pulse ratios than ChR2-evoked EPSPs and were mildly facilitating at short (20ms) intervals (Fig. 3B-C, open circles: n = 5 cells, 3 mice), consistent with an octopus cell’s ability to respond reliably to click trains in vivo5,7,20,48. In contrast, previous results using electrical stimulation demonstrated short-term depression of SGN inputs to octopus cells49,50. However, these experiments were carried out in the presence of higher, non-physiological levels of extracellular calcium. We repeated paired pulse plasticity experiments with non-physiological calcium concentrations (2.4mM) and similarly found that electrically-evoked EP-SPs from SGNs resulted in short-term depression at 50Hz of electrical stimulation (Fig. 3C, grey: n = 3 cells, 2 mice), though not to the degree observed when using full-field, ChR2-evoked inputs.

Glycine evokes inhibitory post synaptic potentials that are occluded by a low input resistance

Given the increased density of inhibitory synapses on octopus cell dendrites, we considered the possibility that somatic and dendritic compartments contribute differently to the temporal computation made by octopus cells. A role for inhibition has never been incorporated into octopus cell models as previous efforts failed to reveal physiological evidence of functional inhibitory synapses onto octopus cells either in vitro12,51,52 or in vivo29. Similarly, we did not observe light-evoked (Glyt2-ChR2) inhibitory post synaptic potentials (IPSPs) in octopus cell somas during whole-cell current clamp recordings from P30-45 mice. Since inhibitory synapses are located primarily on octopus cell dendrites, we posited that their voltage spread to the soma is limited given the extremely low input resistance of octopus cell somas.

To decrease electrotonic isolation of the dendrites and increase input resistance, we pharmacologically blocked voltagegated potassium (Kv) and hyperpolarization-activated cyclic nucleotide-gated (HCN) channels using 100µM 4-Aminopyridine (4-AP) and 50µM ZD 7288 (ZD). This cocktail increased octopus cell membrane resistance (Fig. 4A). To isolate inhibition, 15µM 2,3-dioxo-6-nitro-7-sulfamoyl-benzo[f]quinoxaline (NBQX) was added to block AMPA receptor activation. Consistent with our hypothesis, the increase in input resistance unveiled light-evoked IPSPs in recordings from octopus cell somas (Fig. 4B). Bath application of 500nM strychnine (STN) fully abolished IPSPs (Fig. 4C), confirming functional glycinergic inhibitory synaptic transmission onto octopus cells.

Octopus cells receive glycinergic inhibitory post synaptic potentials.

(A) Voltage responses to a -200pA current injection. This representative neuron hyperpolarized 0.7mV (black) in control conditions. After bath application of 100µM 4-Aminopyridine (4-AP) and 50µM ZD 7288 (ZD), hyperpolarizing responses to the same -200pA current injection increased to 8.8mV at steady state (green). (B) Postsynaptic responses to ChR2 stimulation of glycinergic terminals (Glyt2-ChR2) with a 5Hz train (gray) of 1ms full-field blue light pulses before (black) and after bath application of 100µM 4-AP, 50µM ZD, and 15µM NBQX (green: n = 9 cells, 8 mice). Increased input resistance reveals inhibitory potentials that are difficult to detect. (C) Postsynaptic responses to Glyt2-ChR2 stimulation after bath application of 100µM 4-AP, 50µM ZD, and 15µM NBQX (green), with further sequential addition of 20µM picrotoxin (PTX, pink), 100µM cyclothiazide (CTZ, blue), and 500nM strychnine (STN, orange; n = 6 cells, 5 mice). (D) Change in membrane voltage in response to hyperpolarizing somatic current steps in a morphologically and biophysically realistic model of octopus cells before (black) and after removal of voltage-gated potassium (Kv) and hyperpolarization-activated cyclic nucleotide-gated (HCN) channels (blue). As in in vitro current-clamp recordings, removing Kv and HCN channels increased the magnitude of voltage responses (ΔVm) to hyperpolarizing current. (E) IPSPs measured at the soma of a modeled octopus cell before (black) and after removal of Kv and HCN channels (blue). As in in vitro current-clamp recordings, this allows for somatic IPSP detection. (F) IPSP magnitude in experimental data (green) and the model (blues) as a function of input resistance. In somatic measurements, IPSP size increases with input resistance. Modeled IPSPs are shown for two conductance levels (1nS, dark blue; 100nS light blue). (G) Fold change in magnitude of soma-measured IPSPs (blue) or dendrite-measured IPSCs at proximal (dark orange) and distal (light orange) dendritic locations after removal of Kv and HCN channels. Kv and HCN block increases the magnitude of soma-measured IPSPs. The size of dendritic IPSCs are not changed with Kv and HCN or synapse location.

To determine the types of glycinergic receptors contributing to IPSPs, we pharmacologically blocked subsets of glycine receptors (Fig. 4C). IPSPs were reduced upon addition of 20µM picrotoxin (PTX), which blocks homomeric glycine receptors53–55. Sequential addition of 100µM cyclothiazide (CTZ), which blocks α2-containing homomeric and heteromeric glycine receptors56,57, nearly abolished the remaining IPSPs, and responses were fully abolished with further application of 500nM STN. Collectively, these data demonstrate that glycinergic synaptic contacts onto octopus cell dendrites are functional. Additionally, glycine receptor subunit composition (Fig. 4C) implicates a role for both large conductance extrasynaptic β-subunit lacking homomeric receptors and synaptically localized α2β receptors with slower kinetics58–60.

To confirm whether the confinement of IPSPs to the dendrites is consistent with our understanding of octopus cell biophysics, we developed an improved biophysically and anatomically accurate model of octopus cells based on our findings (Supp. Fig. 4)18,61. This model performed as predicted based on our experimental results. As in our current-clamp recordings (Fig. 4A), removal of Kv and HCN conductances in the model changed the input resistance and current-voltage relationship of the neuron, resulting in reduced electronic isolation (Fig. 4D, blue). In control conditions, stimulation of dendritic glycinergic conductances induced negligible hyperpolarizing voltage changes (Fig. 4E, black). With increased input resistance, dendritic IPSPs measured at the soma were similar to in vitro recordings (Fig. 4E-F). Thus, this model confirms that dendritic IPSPs can elicit somatic hyperpolarization in octopus cells when electrotonic dendritic isolation is reduced.

While blocking Kv and HCN allowed us to reveal IPSPs at the soma, 4-AP increases the duration of the already unphysiological ChR2-evoked presynaptic action potential47, resulting in altered release probabilities and synaptic properties, amongst other caveats62. To directly confirm that the increase in somatically-measured IPSP amplitude can be explained by changes in input resistance alone, we used the octopus cell model to simulate dendritic glycinergic conductances and measure changes in dendritic current and somatic potential amplitude in the presence of blocked Kv and HCN channels. Kv and HCN block and the resulting change in input resistance increased the magnitude of soma-measured IPSPs for all glycine conductances (Fig. 4G, blue). Dendrite-measured currents did not change as a result of the hyperpolarization induced by Kv and HCN block regardless of synaptic location on the dendrites or the magnitude of the glycine conductance (Fig 4G, dark orange, light orange). Together, results from the model replicate those collected in vitro and provide evidence of functional glycinergic synaptic transmission that is difficult to detect with in vitro somatic recordings.

Inhibition decreases the magnitude and advances the timing of dendritic SGN inputs

SGN synapses onto octopus cell dendrites are arranged tonotopically, with higher frequency SGNs from the base of the cochlea terminating on the distal dendrites and lower frequency SGNs from more apical positions terminating more proximally (Fig. 1A). This organization has been proposed to re-synchronize coincidentally firing SGNs that are activated at slightly different times due to the time it takes for the sound stimulus to travel from the base to the apex of the cochlea18. To test how inhibition in the dendrites shapes coincidence detection, we first used our model to explore the influence of simultaneous activation of inhibitory and excitatory synapses at varying locations along the dendritic tree. By placing inhibitory synapses on proximal or distal dendrites and moving the relative location of excitation, we modelled the effect of on-path and off-path inhibition63 on somatically recorded EPSPs (Fig. 5A, Supp. Fig. 5). For both inhibition proximal to excitation (Fig. 5B-C: on-path) and inhibition distal to excitation (Fig. 5D-E: off-path), the model predicted that inhibition reduces EPSP amplitude and accelerates EPSP peak timing at the soma. Thus, the presence of inhibition appears to enable modulation of EPSP timing in dendritic compartments during continuous auditory stimuli when inhibition can be recruited beyond the onset of a sound and thus allow for adaptable temporal processing during the duration of a stimulus.

Coincident excitation and inhibition on octopus cell dendrites advances EPSP peak times.

(A) The impact of distance between excitatory and inhibitory synapses was measured in a computational model of octopus cells. Inhibitory synapses were placed either proximally (on-path) or distally (off-path) to excitation. Excitatory synapses were placed at varying locations along the dendritic arbor to change the anatomical distance (Δd) where d=0 is the location of inhibition (GGly) and d=1 is the condition where excitation (GAMPA) and inhibition are maximally separated. EPSPs were measured at the soma in all conditions (green). (B-E) Quantification of the percent change in soma-measured EPSP magnitude and the shift in EPSP peak timing in models of (B-C) on-path and (D-E) off-path inhibition. Example traces show EPSPs with (green) and without (black) inhibition at d=0 and d=1. Distal dendrites (d=1) have higher local input resistance and lower IPSP attenuation due to the sealed end. Inset scale bars are 1mV, 200ms. (F-H) Coincident stimulation of excitation and inhibition changes EPSP shape. (F) Representative responses to independent stimulation of excitatory SGNs (black), independent stimulation of inhibitory inputs (light blue), coincident stimulation of both excitation and inhibition (green), and independent stimulation of excitatory SGNs with the addition of 1µM strychnine (STN). Quantification of (G) the percent change in EPSP height and (H) the shift in EPSP peak timing during coincident Glyt2-ChR2 activation of inhibitory inputs (green: n = 8 cells, 6 mice), bath application of 25µM glycine (dark blue: n = 4 cells, 3 mice), and bath application of 1µM STN (orange: n = 5 cells, 4 mice). Activation of glycinergic receptors during excitation decreases EPSP heights and advances EPSP peaks. Blocking of tonically active glycine receptors slows and delays EPSPs. Data are presented as mean ± SD. Markers represent the average quantification for a cell.

To directly test if the model’s prediction that temporally coincident excitation and inhibition affects the timing and amplitude of excitatory SGN inputs as they travel towards the octopus cell soma, we coincidently activated SGNs and glycinergic inputs in vitro. In these experiments, the octopus cell properties were not altered pharmacologically and inhibition was undetectable or only visible with averaging over many sweeps (Fig. 5F, blue). When synaptic inhibition was evoked together with excitation (Fig. 5F, green), the soma-recorded EPSPs were smaller than when excitation was evoked alone (Fig. 5F, black: n = 8 cells, 6 mice). ChR2-evoked inhibition decreased EPSP heights by 25.2 ± 9.0% (Fig. 5G, green) and shifted the peak of EPSPs forward 57.5 ± 26µs (Fig. 5H, green). This effect was mimicked by bath application of 25µM glycine (Fig. 5G-H, blue: n = 4 cells, 3 mice). Further, bath application of 1µM STN had the opposite effect, resulting in larger EPSPs, delayed peak times, and increased half-widths (Fig. 5F-H, orange: n = 5 cells, 4 mice). Thus, the timing of EPSP arrival may be shaped both by the release of synaptic glycine and by tonically active glycine channels. Of note, many SGNs also terminate on the octopus cell soma, where inhibition is minimal. This suggests that the octopus cell’s ability to act as a coincidence detector depends on two stages of compartmentalized computations, one in the dendrite that combines excitation and inhibition to provide important information about which frequencies co-occur in a complex sound stimulus and one in the soma that is restricted by the rigid temporal summation window for coincidence detection. Together with the electrotonic properties of the octopus cell and the dominance of low threshold, low jitter Ia SGN inputs, these combined computations can enable reliable coincidence detection and proper cross-frequency binding needed for perception of sound.

Discussion

Coincidence detection plays a critical role in many cognitive and perceptual processes, from the ability to localize sound, to the binding of auditory and visual features of a common stimulus. Depending on the computation, the temporal window for integration can range widely, thereby requiring circuitry with distinct anatomical and physiological properties. Here, we describe a two-domain mechanism for coincidence detection that can detect co-occurring frequencies with different degrees of precision. By mapping and selectively activating synaptic inputs onto octopus cells both in vitro and in a computational model, we revealed that compartmentalized dendritic nonlinearities impact the temporal integration window under which somatic coincidence detection computations are made. The arrival of many small, reliable excitatory inputs (Fig. 3) from low-threshold SGNs (Fig. 2) is continuous throughout an ongoing stimulus. We demonstrate that glycinergic inhibition to octopus cell dendrites (Fig. 1) can shift the magnitude and timing of SGN EPSPs as they summate in the soma (Fig. 4,5). The narrow window for coincidence detection computations allows the octopus cell to respond with temporal precision using momentary evidence provided by SGNs at the onset of the stimulus. We propose that, as a stimulus persists, inhibition onto octopus cell dendrites can adjust the timing of excitation before arriving at the soma for the final input-output computation. This allows the cell to make an additional computation with a slightly longer window for evidence accumulation without compromising the accuracy of the somatic onset computation (Fig. 6).

Proposed model of flexible dendritic timing for precise somatic coincidence detection.

(A-B) Co-occurring frequencies in broadband onsets and frequency modulations occur on different time scales. (Top) Spectrograms illustrate frequency strength over time. (Bottom) Waveform illustrate total intensity over time. Both of these stimuli drive firing in octopus cells despite the limitations imposed by a ∼1ms window for temporal integration. (C) Excitation must summate at the soma of octopus cells within a narrow time window (∼1ms) to achieve a depolarization rate rapid enough to trigger action potentials. Summation of excitation alone accounts for responses to sound onsets (inset, top). Stimuli that require summation over a longer time period, such as frequency modulated sweeps, require synaptic inhibition to modulate the timing of excitation on octopus cell dendrites before they reach the soma for coincidence detection computations (inset, bottom). We propose a mechanism for preferential processing of a subset of excitatory inputs where selective temporal advancement of a subset of EPSPs could expand the effective window for coincidence detection at the soma.

As coincidence detectors in the auditory system, octopus cells are faced with the challenge of recognizing complex sounds that include many frequencies that co-occur from the beginning to the end of the stimulus. As shown by in vivo recordings6,29, octopus cells respond well to cues that include complex spectrotemporal patterns, including frequency modulations beyond the onset of the stimulus29. Given that the auditory environment is filled with overlapping sound stimuli, such responses presumably allow the octopus cell to encode which frequencies belong to which sound. Our data thus support the role of octopus cells beyond simple onset coincidence detectors that rely solely on the temporal summation of excitation. The results suggest that, despite high Kv and HCN conductances at rest, the addition of dendritic inhibition transforms the magnitude and timing of SGN signals as they arrive in the cell body, which may expand their response selectivity and allow them to become a slightly leakier integrator and thus accumulate evidence beyond onsets. Although, this inhibition is difficult to detect because of shunting, our data demonstrate that it is both present and impactful.

As well as needing to work beyond onsets, an effective coincidence detector in the auditory system must also function reliably across a range of sound intensities. Intensity information is encoded by the number and types of SGNs that are activated in the cochlea. The Ia, Ib, and Ic molecular subtypes defined in mouse40,41,43 broadly correspond to the anatomically and physiologically defined subtypes described across species42,64. We find that the majority of inputs onto octopus cells come from Ia SGNs, which most closely correspond to the low-threshold, high-spontaneous rate population. Consistent with this result, single-unit SGN recordings in cats demonstrated a bias towards low-threshold, high-spontaneous rate axon collaterals in the octopus cell area33. Low-threshold SGNs are also characterized by short first spike latencies and low temporal jitter65–68. A hallmark of the octopus cell is the fact that it only fires action potentials when many SGN inputs are activated within a narrow period of time69. The presence of many low-threshold and temporally precise inputs on the octopus cell may help ensure that coincidence detection still works reliably for quiet sounds. Further, Ia inputs onto octopus cells do not exhibit paired-pulse depression, not unlike low levels of depression seen in Ia inputs to bushy cells70. The presence of SGN inputs without paired-pulse depression could be beneficial for encoding sustained auditory signals. Finally, although Ia SGNs are over-represented, Ib and Ic inputs are also present. Since precise, low-threshold SGN responses can be saturated by background noise such that responses to relevant stimuli are masked38,65,68,71, recruitment of higher threshold SGNs at higher sound intensities may compensate for this tradeoff.

The presence of inhibitory inputs onto dendrites is a fundamental feature of the nervous system and, in other systems, contributes to a neuron’s final computation. For example, direction selectivity computations in dendrites of retinal cells require excitation-inhibition interactions in dendritic compartments72,73. In pyramidal cells of the cortex and hippocampus, the spatial distribution of inhibition impacts dendritic non-linearities in a branch selective manner74–79. However, octopus cells do not share all mechanisms for dendritic computation as neurons in the cortex and hippocampus. The lack of backpropagating action potentials and dendritic calcium spikes in octopus cells causes submillisecond timing differences in dendritic summation of coincident excitation and local inhibition to be consequential in subthreshold computation before somatic action potential generation.

Although this work uncovers a role for inhibition, understanding of octopus cell computations is limited by the fact that it remains unclear what kind of information is carried by inhibitory inputs. Although the presence of presynaptic glycinergic puncta in the octopus cell area80–83 and glycinergic receptor expression in octopus cells84–88 is well established, it is unknown where this glycinergic innervation originates. Of the local neurons within the CNC that provide inhibition to the VCN, there is no evidence of connections to octopus cells from D-Stellate52, L-Stellate89, or tuberculoventral cells90. Outside of the CNC, terminal degeneration experiments in cats suggested the superior periolivary nucleus (SPON) and the ventral division of the lateral lemniscus (VNLL) as potential sources of descending inhibition to the octopus cell area91. Octopus cells provide excitatory input to both the SPON92–96 and the VNLL20,97–101, raising the possibility of feedback inhibition from the auditory brainstem as a circuit mechanism for elongated temporal summation windows during ongoing stimuli. Although feedback inhibition is not rapid enough to prevent or alter the onset response that octopus cells are well-known for, it could limit the duration of a response or change the effective coincidence detection window as the stimulus continues. Future studies will be required to identify the source of inhibition and its organization along the dendrites. If inhibitory inputs tonotopically match the local, narrowly-tuned dendritic SGN inputs, it is possible that frequency matched inhibition could influence spectral selectivity or feature extraction. On the other hand, broadly tuned inhibition could reduce depolarization block or serve as a temporal milestone that signals gaps or offsets. Further characterization of in vivo octopus cell responses in complex sound environments may clarify the effect of noise on signal detection and could reveal additional features of this cell’s contributions to perception of the auditory world.

eLife Assessment

This valuable work analyzes how specialized cells in the auditory cells, known as the octopus cells, can detect coincidences in their inputs at the submillisecond time scale. While previous work indicated that these cells receive no inhibitory inputs, the present study unambiguously demonstrates that these cells receive inhibitory glycinergic inputs. The physiologic impact of these inputs needs to be studied further. It remains incomplete at present but could be improved by addressing caveats related to similar sizes of excitatory postsynaptic potentials and spikes in the octopus neurons.

Reviewer #1 (Public Review):

Kreeger and colleagues have explored the balance of excitation and inhibition in the cochlear nucleus octopus cells of mice using morphological, electrophysiological, and computational methods. On the surface, the conclusion, that synaptic inhibition is present, does not seem like a leap. However, the octopus cells have been in the past portrayed as devoid of inhibition. This view was supported by the seeming lack of glycinergic fibers in the octopus cell area and the lack of apparent IPSPs. Here, Kreeger et al. used beautiful immunohistochemical and mouse genetic methods to quantify the inhibitory and excitatory boutons over the complete surface of individual octopus cells and further analysed the proportions of the different subtypes of spiral ganglion cell inputs. I think the analysis stands as one of the most complete descriptions of any neuron, leaving little doubt about the presence of glycinergic boutons.

Kreeger et al then examined inhibition physiologically, but here I felt that the study was incomplete. Specifically, no attempt was made to assess the actual, biological values of synaptic conductance for AMPAR and GlyR. Thus, we don't really know how potent the GlyR could be in mediating inhibition. Here are some numbered comments:

(1) "EPSPs" were evoked either optogenetically or with electrical stimulation. The resulting depolarizations are interpreted to be EPSPs. However previous studies from Oertel show that octopus cells have tiny spikes, and distinguishing them from EPSPs is tricky. No mention is made here about how or whether that was done. Thus, the analysis of EPSP amplitude is ambiguous.

(2) For this and later analysis, a voltage clamp of synaptic inputs would have been a simple alternative to avoid contaminating spikes or shunts by background or voltage-gated conductances. Yet only the current clamp was employed. I can understand that the authors might feel that the voltage clamp is 'flawed' because of the failure to clamp dendrites. But that may have been a good price to pay in this case. The authors should have at least justified their choice of method and detailed its caveats.

(3) The modeling raised several concerns. First, there is little presentation of assumptions, and of course, a model is entirely about its assumptions. For example, what excitatory conductance amplitudes were used? The same for inhibitory conductance? How were these values arrived at? The authors note that EPSGs and IPSGs had peaks at 0.3 and 3 ms. On what basis were these numbers obtained? The model's conclusions entirely depend on these values, and no measurements were made here that could have provided them. Parenthetical reference is made to Figure S5 where a range of values are tested, but with little explanation or justification.

(4) In experiments that combined E and I stimulation, what exactly were time timecourses of the conductance changes, and how 'synchronous' were they, given the different methods to evoke them? (had the authors done voltage clamp they would know the answers).

(5) Figure 4G is confusing to me. Its point, according to the text, is to show that changes in membrane properties induced by a block of Kv and HCN channels would not be expected to alter the amplitudes of EPSCs and IPSCs across the dendritic expanse. Now we are talking about currents (not shunting effects), and the presumption is that the blockers would alter the resting potential and thus the driving force for the currents. But what was the measured membrane potential change in the blockers? Surely that was documented. To me, the bigger concern (stated in the text) is whether the blockers altered exocytosis, and thus the increase in IPSP amplitude in blockers is due BOTH to loss of shunting and increase in presynaptic spike width. Added to this is that 4AP will reduce the spike threshold, thus allowing more ChR2-expressing axons to reach the threshold. Figure 4G does not address this point.

(6) Figure 5F is striking as the key piece of biological data that shows that inhibition does reduce the amplitude of "EPSPs" in octopus cells. Given the other uncertainties mentioned, I wondered if it makes sense as an example of shunting inhibition. Specifically, what are the relative synaptic conductances, and would you predict a 25% reduction given the actual (not modeled) values?

(7) Some of the supplemental figures, like 4 and 5, are hardly mentioned. Few will glean anything from them unless the authors direct attention to them and explain them better. In general, the readers would benefit from more complete explanations of what was done.

Reviewer #2 (Public Review):

Summary:

Kreeger et.al provided mechanistic evidence for flexible coincidence detection of auditory nerve synaptic inputs by octopus cells in the mouse cochlear nucleus. The octopus cells are specialized neurons that can fire repetitively at very high rates (> 800 Hz in vivo), yield responses dominated by the onset of sound for simple stimuli, and integrate auditory nerve inputs over a wide frequency span. Previously, it was thought that octopus cells received little inhibitory input, and their integration of auditory input depended principally on temporally precise coincidence detection of excitatory auditory nerve inputs, coupled with a low input resistance established by high levels of expression of certain potassium channels and hyperpolarization-activated channels.

In this study, the authors used a combination of numerous genetic mouse models to characterize synaptic inputs and enable optogenetic stimulation of subsets of afferents, fluorescent microscopy, detailed reconstructions of the location of inhibitory synapses on the soma and dendrites of octopus cells, and computational modeling, to explore the importance of inhibitory inputs to the cells. They determined through assessment of excitatory and inhibitory synaptic densities that spiral ganglion neuron synapses are densest on the soma and proximal dendrite, while glycenergic inhibitory synaptic density is greater on the dendrites compared to the soma of octopus cells. Using different genetic lines, the authors further elucidated that the majority of excitatory synapses on the octopus cells are from type 1a spiral ganglion neurons, which have low response thresholds and high rates of spontaneous activity. In the second half of the paper, the authors employed electrophysiology to uncover the physiological response of octopus cells to excitatory and inhibitory inputs. Using a combination of pharmacological blockers in vitro cellular and computational modeling, the authors conclude that glycine in fact evokes IPSPs in octopus cells; these IPSPs are largely shunted by the high membrane conductance of the cells under normal conditions and thus were not clearly evident in prior studies. Pharmacological experiments point towards a specific glycine receptor subunit composition. Lastly, Kreeger et. al demonstrated with in vitro recordings and computational modeling that octopus cell inhibition modulates the amplitude and timing of dendritic spiral ganglion inputs to octopus cells, allowing for flexible coincidence detection.

Strengths:

The work combines a number of approaches and complementary observations to characterize the spatial patterns of excitatory and inhibitory synaptic input, and the type of auditory nerve input to the octopus cells. The combination of multiple mouse lines enables a better understanding of and helps to define, the pattern of synaptic convergence onto these cells. The electrophysiology provides excellent functional evidence for the presence of the inhibitory inputs, and the modeling helps to interpret the likely functional role of inhibition. The work is technically well done and adds an interesting dimension related to the processing of sound by these neurons. The paper is overall well written, the experimental tests are well-motivated and easy to follow. The discussion is reasonable and touches on both the potential implications of the work as well as some caveats.

Weaknesses:

While the conclusions presented by the authors are solid, a prominent question remains regarding the source of the glycinergic input onto octopus cells. In the discussion, the authors claim that there is no evidence for D-stellate, L-stellate, and tuberculoventral cell (all local inhibitory neurons of the ventral and dorsal cochlear nucleus) connections to octopus cells, and cite the relevant literature. An experimental approach will be necessary to properly rule out (or rule in) these cell types and others that may arise from other auditory brainstem nuclei. Understanding which cells provide the inhibitory input will be an essential step in clarifying its roles in the processing of sound by octopus cells.

The authors showed that type 1a SGNs are the most abundant inputs to octopus cells via microscopy. However, in Figure 3 they compare optical stimulation of all classes of ANFs, then compare this against stimulation of type 1b/c ANFs. While a difference in the paired-pulse ratio (and therefore, likely release probability) can be inferred by the difference between Foxg1-ChR2 and Ntng1-ChR2, it would have been preferable to have specific data with selective stimulation of type 1a neurons.

Author response:

The following is the authors’ response to the original reviews.

eLife assessment

This valuable work analyzes how specialized cells in the auditory cells, known as the octopus cells, can detect coincidences in their inputs at the submillisecond time scale. While previous work indicated that these cells receive no inhibitory inputs, the present study unambiguously demonstrates that these cells receive inhibitory glycinergic inputs. The physiologic impact of these inputs needs to be studied further. It remains incomplete at present but could be made solid by addressing caveats related to similar sizes of excitatory postsynaptic potentials and spikes in the octopus neurons.

We apologize for not explicitly describing our experimental methods and analyses procedures that ensure the discrimination between action potentials and EPSPs. This has been addressed in responses to reviewer comments and amended in the manuscript.

Reviewer #1 (Public Review):

Kreeger and colleagues have explored the balance of excitation and inhibition in the cochlear nucleus octopus cells of mice using morphological, electrophysiological, and computational methods. On the surface, the conclusion, that synaptic inhibition is present, does not seem like a leap. However, the octopus cells have been in the past portrayed as devoid of inhibition. This view was supported by the seeming lack of glycinergic fibers in the octopus cell area and the lack of apparent IPSPs. Here, Kreeger et al. used beautiful immunohistochemical and mouse genetic methods to quantify the inhibitory and excitatory boutons over the complete surface of individual octopus cells and further analyzed the proportions of the different subtypes of spiral ganglion cell inputs. I think the analysis stands as one of the most complete descriptions of any neuron, leaving little doubt about the presence of glycinergic boutons.

Kreeger et al then examined inhibition physiologically, but here I felt that the study was incomplete. Specifically, no attempt was made to assess the actual, biological values of synaptic conductance for AMPAR and GlyR. Thus, we don't really know how potent the GlyR could be in mediating inhibition. Here are some numbered comments:

(1) "EPSPs" were evoked either optogenetically or with electrical stimulation. The resulting depolarizations are interpreted to be EPSPs. However previous studies from Oertel show that octopus cells have tiny spikes, and distinguishing them from EPSPs is tricky. No mention is made here about how or whether that was done. Thus, the analysis of EPSP amplitude is ambiguous.

We agree that large EPSPs can be difficult to distinguish from an octopus cell’s short spikes during experiments. During analysis, we distinguished spikes from EPSPs by generating phase plots, which allow us to visualize the first derivative of the voltage trace on the y-axis and the value of the voltage on the x-axis at each moment in time. In the example shown below, four depolarizing events were electrically evoked in an octopus cell (panel A). The largest of these events (shown in orange in panels B-D) has an amplitude of ~9mV and could be a small spike. The first derivative of the voltage (panel C) reveals a bi-phasic response in the larger orange trace, where during the rising phase (mV/ms > 0) of the EPSP there is a second, sharper rising phase for the spike. Like more traditionally sized action potentials, phase plots for octopus cell spikes also reveal a sharp change in the rate of voltage change over time (Author response image 1 panel D, ✱) after the rising action of the EPSP begins to slow. EPSPs (shown in blue in panels B-D) lack the deflection in the phase plot. Not all cases were as unambiguous as this example. Therefore, our analysis only included subthreshold stimulation that unambiguously evoked EPSPs, not spikes. A brief description of this analysis has been added to the methods text (lines 625-627) and we have noted in the results section that both ChR2-evoked and electrically-evoked stimulation can produce small action potentials, which were excluded from analysis (lines 156-158).

(2) For this and later analysis, a voltage clamp of synaptic inputs would have been a simple alternative to avoid contaminating spikes or shunts by background or voltage-gated conductances. Yet only the current clamp was employed. I can understand that the authors might feel that the voltage clamp is 'flawed' because of the failure to clamp dendrites. But that may have been a good price to pay in this case. The authors should have at least justified their choice of method and detailed its caveats.

We agree that data collected using voltage-clamp would have eliminated the confound of short action potentials and avoided the influence of voltage-gated conductances. The large-diameter, and comparatively simple dendritic trees of octopus cells make them good morphological candidates for reliable voltage clamp. However, as suggested, we were concerned that the abundance of channels open at the neuron’s resting potential would make it difficult to sufficiently clamp dendrites. Ultimately, given the low input resistances of octopus cells and the fast kinetics of excitatory inputs, we determined that bad voltage clamp conditions were likely to result in unclamped synaptic events with unpredicted distortions in kinetics and attenuation (To et al. 2022; PMID: 34480986; DOI: 10.1016/j.neuroscience.2021.08.024). We therefore chose to focus our efforts on current-clamp.

Beyond the limits of both current-clamp and voltage-clamp, we chose to leave all conductances that influence EPSP dendritic propagation intact because our model demonstrates that active Kv and leak conductances shape and attenuate synaptic inputs as they travel through the dendritic tree (Supp. Fig. 4F-G). The addition of voltage-clamp recordings would not impact the conclusions we make about EPSP summation at the soma. Future studies will need to focus on a dendrite-centric view of local excitatory and inhibitory summation. For dendrite-centric experiments, dendritic voltage-clamp recordings are well suited to answer that set of questions.

(3) The modeling raised several concerns. First, there is little presentation of assumptions, and of course, a model is entirely about its assumptions. For example, what excitatory conductance amplitudes were used? The same for inhibitory conductance? How were these values arrived at? The authors note that EPSGs and IPSGs had peaks at 0.3 and 3 ms. On what basis were these numbers obtained? The model's conclusions entirely depend on these values, and no measurements were made here that could have provided them. Parenthetical reference is made to Figure S5 where a range of values are tested, but with little explanation or justification.

We apologize for not providing this information. We used our octopus neuron model to fit both EPSP and IPSP parameters to match experimental data. We have expanded the methods to include final values for the conductances (lines 649-651), which were adjusted to match experimental values seen in current-clamp recordings. We have also expanded the results section to describe each of the parameters we tuned (lines 203-222). An example of these adjustments is illustrated in Fig. 4F where the magnitude of inhibitory potentials at different conductances (100nS and 1nS) was compared to experimental data over a range of octopus cell input resistance conditions. Kinetic parameters were determined by aligning modeled PSPs to the rise times and full width at half maximum (FWHM) measurements from experiments under control and Kv block conditions. The experimental data for EPSPs and IPSPs that was used to fit the model is shown in Author response image 2 below.

(4) In experiments that combined E and I stimulation, what exactly were time courses of the conductance changes, and how 'synchronous' were they, given the different methods to evoke them? (had the authors done voltage clamp they would know the answers).

We chose to focus data collection on voltage changes at the soma under physiological conditions to better understand how excitation and inhibition integrate at the somatic compartment. Our conclusions in the combined E and I stimulation experiments require the resting membrane properties of octopus cells to be intact to make physiologically-relevant conclusions. Our current-clamp data includes the critical impact of leak, Kv, and HCN conductances on this computation. Reliable voltage-clamp would necessitate the removal of the Kv and HCN conductances that shape PSP magnitude, shape, and speed. Because it was not necessary to measure the conductances and kinetics of specific channels, we chose to use current-clamp.

Evoked IPSPs and EPSPs had cell-to-cell variability in their latencies to onset. Somatically-recorded optically-evoked inhibition under pharmacological conditions that changed cable properties had onset latencies between 2.5 and 4.3ms; electrically-evoked excitation under control conditions had latencies between 0.8 and 1.4ms. To overcome cell-to-cell timing variabilities, we presented a shuffled set of stimulation pairings that had a 3ms range of timings with 200µs intervals. As the evoked excitation and inhibition become more ‘synchronous’, the impact on EPSP magnitude and timing is greatest. Data presented in this paper was for the stimulation pairings that evokes a maximal shift in EPSP timing. On average, this occurred when the optical stimulation began ~1.2ms before electrical stimulation. Stimulation pairing times ranged between a 0ms offset and a 1.8ms offset at the extremes. An example of the shuffled stimulation pairings is shown in Author response image 3 below, and we have included information about the shuffled stimulus in the methods (lines 627-630)

(5) Figure 4G is confusing to me. Its point, according to the text, is to show that changes in membrane properties induced by a block of Kv and HCN channels would not be expected to alter the amplitudes of EPSCs and IPSCs across the dendritic expanse. Now we are talking about currents (not shunting effects), and the presumption is that the blockers would alter the resting potential and thus the driving force for the currents. But what was the measured membrane potential change in the blockers? Surely that was documented. To me, the bigger concern (stated in the text) is whether the blockers altered exocytosis, and thus the increase in IPSP amplitude in blockers is due BOTH to loss of shunting and increase in presynaptic spike width. Added to this is that 4AP will reduce the spike threshold, thus allowing more ChR2-expressing axons to reach the threshold. Figure 4G does not address this point.

These are valuable points that motivated us to improve the clarity of this figure and the corresponding text. We discussed two separate points in this paragraph and were not clear. Our intention with Figure 4G was to address concerns that using pharmacological blockers changes driving forces and may confound the measured change in magnitude of postsynaptic potentials. Membrane potentials hyperpolarized by approximately 8-10 mV after application of blockers. We corrected for this effect by adding a holding current to depolarize the neuron to its baseline resting potential. Text in the results (lines 187-190) and figure legends have been changed to clarify these points.

We also removed any discussion of presynaptic effects from this portion of the text because our description was incomplete and we did not directly collect data related to these claims. We originally wrote, “While blocking Kv and HCN allowed us to reveal IPSPs at the soma, 4-AP increases the duration of the already unphysiological ChR2-evoked presynaptic action potential (Jackman et al., 2014; DOI: 10.1523/jneurosci.4694-13.2014), resulting in altered release probabilities and synaptic properties, amongst other caveats (Mathie et al., 1998; DOI: 10.1016/S0306-3623(97)00034-7)”. Ultimately, effects on exocytosis, presynaptic excitability, or release probability are only relevant for the experiments presented in Figure 4. Figure 4 serves as evidence that synaptic release of glycine elicits strychnine-sensitive inhibitory postsynaptic potentials in octopus cells. Concerns of presynaptic effects do not carry over to the data presented in Figure 5, as Kv and HCN were not blocked in these experiments. Therefore, we have removed this portion of the text.

(6) Figure 5F is striking as the key piece of biological data that shows that inhibition does reduce the amplitude of "EPSPs" in octopus cells. Given the other uncertainties mentioned, I wondered if it makes sense as an example of shunting inhibition. Specifically, what are the relative synaptic conductances, and would you predict a 25% reduction given the actual (not modeled) values?

We agree that both shunting and hyperpolarizing inhibition could play a role in the measured EPSP changes. Because we focused data collection on voltage changes at the soma under physiological conditions, we cannot calculate the relative synaptic conductances. Together, our experimental current-clamp results paired with estimates from the model provide compelling evidence for the change we observe in EPSPs. Regardless, the relative weights of the synaptic conductances is a very interesting question, but this information is not necessary to answer the questions posed in this study, namely the impact of dendritic inhibition on the arrival of EPSPs in the soma.

(7) Some of the supplemental figures, like 4 and 5, are hardly mentioned. Few will glean anything from them unless the authors direct attention to them and explain them better. In general, the readers would benefit from more complete explanations of what was done.

We apologize for not fully discussing these figures in the results text. We have fully expanded the results section to detail the experiments and results presented in the supplement (lines 203-238).

Reviewer #2 (Public Review):

Summary:

Kreeger et.al provided mechanistic evidence for flexible coincidence detection of auditory nerve synaptic inputs by octopus cells in the mouse cochlear nucleus. The octopus cells are specialized neurons that can fire repetitively at very high rates (> 800 Hz in vivo), yield responses dominated by the onset of sound for simple stimuli, and integrate auditory nerve inputs over a wide frequency span. Previously, it was thought that octopus cells received little inhibitory input, and their integration of auditory input depended principally on temporally precise coincidence detection of excitatory auditory nerve inputs, coupled with a low input resistance established by high levels of expression of certain potassium channels and hyperpolarization-activated channels.

In this study, the authors used a combination of numerous genetic mouse models to characterize synaptic inputs and enable optogenetic stimulation of subsets of afferents, fluorescent microscopy, detailed reconstructions of the location of inhibitory synapses on the soma and dendrites of octopus cells, and computational modeling, to explore the importance of inhibitory inputs to the cells. They determined through assessment of excitatory and inhibitory synaptic densities that spiral ganglion neuron synapses are densest on the soma and proximal dendrite, while glycinergic inhibitory synaptic density is greater on the dendrites compared to the soma of octopus cells. Using different genetic lines, the authors further elucidated that the majority of excitatory synapses on the octopus cells are from type 1a spiral ganglion neurons, which have low response thresholds and high rates of spontaneous activity. In the second half of the paper, the authors employed electrophysiology to uncover the physiological response of octopus cells to excitatory and inhibitory inputs. Using a combination of pharmacological blockers in vitro cellular and computational modeling, the authors conclude that glycine in fact evokes IPSPs in octopus cells; these IPSPs are largely shunted by the high membrane conductance of the cells under normal conditions and thus were not clearly evident in prior studies. Pharmacological experiments point towards a specific glycine receptor subunit composition. Lastly, Kreeger et. al demonstrated with in vitro recordings and computational modeling that octopus cell inhibition modulates the amplitude and timing of dendritic spiral ganglion inputs to octopus cells, allowing for flexible coincidence detection.

Strengths:

The work combines a number of approaches and complementary observations to characterize the spatial patterns of excitatory and inhibitory synaptic input, and the type of auditory nerve input to the octopus cells. The combination of multiple mouse lines enables a better understanding of and helps to define, the pattern of synaptic convergence onto these cells. The electrophysiology provides excellent functional evidence for the presence of the inhibitory inputs, and the modeling helps to interpret the likely functional role of inhibition. The work is technically well done and adds an interesting dimension related to the processing of sound by these neurons. The paper is overall well written, the experimental tests are well-motivated and easy to follow. The discussion is reasonable and touches on both the potential implications of the work as well as some caveats.

Weaknesses:

While the conclusions presented by the authors are solid, a prominent question remains regarding the source of the glycinergic input onto octopus cells. In the discussion, the authors claim that there is no evidence for D-stellate, L-stellate, and tuberculoventral cell (all local inhibitory neurons of the ventral and dorsal cochlear nucleus) connections to octopus cells, and cite the relevant literature. An experimental approach will be necessary to properly rule out (or rule in) these cell types and others that may arise from other auditory brainstem nuclei. Understanding which cells provide the inhibitory input will be an essential step in clarifying its roles in the processing of sound by octopus cells.

We are glad that the reviewer agrees with the conclusions we have made and is interested in learning more about how these findings impact sound processing. We agree that defining the source of inhibition will dramatically shape our understanding of the computation octopus cells are making. However, this is not an easy task, given the small size of the octopus cell area, and will involve considerable additional work. Since the overall findings do not depend on knowing the source of inhibition, we have instead re-written the discussion to clarify the lack of evidence for intrinsic inhibitory inputs to octopus cells, in addition to presenting likely candidates. As genetic profiles of cochlear nucleus and other auditory brainstem neurons become available, we intend to make and utilize genetic mouse models to answer questions like this.

The authors showed that type 1a SGNs are the most abundant inputs to octopus cells via microscopy. However, in Figure 3 they compare optical stimulation of all classes of ANFs, then compare this against stimulation of type 1b/c ANFs. While a difference in the paired-pulse ratio (and therefore, likely release probability) can be inferred by the difference between Foxg1-ChR2 and Ntng1-ChR2, it would have been preferable to have specific data with selective stimulation of type 1a neurons.

We agree that complete genetic access to only the Ia population would have been the preferable approach, but we did not have an appropriate line when beginning these experiments. Because our results did not suggest a meaningful difference between the populations, we did not pursue further investigation once a line was available.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Besides the points mentioned in the main review:

Minor

(1) I really like the graphics and the immunohistological presentation.

(2) Lines 316-319 say that octopus cells lack things like back-propagating spikes and dendritic Ca spikes. How do you know this?

This statement was intended to be a summary of suggestions from the literature and lacked references and context as written. We have rewritten this section and clarified that our hypothesis was formed from data found in the literature (lines 334-337).

(3) Spectrograms of Figure 6A...where were these data obtained?

We recorded and visualized human-generated rhythmic tapping and high-frequency squeaking sounds using Audacity. The visualizations of rhythmic tapping and imitated vocalizations are meant to show two different types of multi-frequency stimuli we hypothesize would result in somatic summation within an octopus cell’s spike integration window, despite differences in timing. We rewrote the figure legend to explain more clearly what is shown and how it relates to the model in Figure 6.

(4) 'on-path' and 'off-path' seem like jargon that may not be clear to the average reader.

Thank you for pointing out our use of unapproachable jargon. We have replaced the term from the figure with “proximal” and “distal” inhibition. In the main text, we now describe on-path and off-path together as the effect of location of dendritic inhibition on somatically recorded EPSPs.

(5) The paper could benefit from a table of modeled values.

We have added specific details about the modelling in the text and clarified which modeled values were referenced from previous computational models and which were tuned to fit experimental data. Since most values were taken from a referenced publication, we did not add a table and instead point readers towards that source.

(6) Figure S4A-C what currents were delivered to the modeled cells?

The model cells were injected with a -0.8 nA DC current for 300 ms in current clamp mode. This information has been added to the figure legend.

(7) In that figure "scaling factors" scale exactly which channels?

Scaling factor is used to scale low-voltage activated K+ (ḡKLT), high threshold K+ (ḡKHT), fast transient K+ (ḡKA), hyperpolarization-activated cyclic nucleotide-gated HCN (ḡh) but not fast Na+ (ḡNa) and leak K+ (ḡleak). This information has been added to the text (lines 205-208 and 646-653).

(8) In performing and modeling Kv/HCN block, do you know how complete the level of the block is?

Since we cannot assess how complete the level of block is, we have changed the language in the text to clarify that we are reducing Kv and HCN channel conductance to the degree needed to increase resistance of the neuron (line 185).

(9) More on this Figure S4. It is hardly referred to in the text except to say that it supports that blocking the Kv/HCN channels will enhance the IPSP. Given how large the figure is, can you offer more of a conclusion than that? Also, in the synaptic model in that figure, the IPSCs are presumably happening in current-clamp conditions, and the reduction in amplitude of the IPSC (as opposed to the increase in IPSP) is due to hyperpolarization. Can you simply state that so readers can track what this figure is showing? Other similar things: what is a transfer impedance? How is it measured? What do we take from the analysis?

We have elaborated on our description of both Supp. Fig. 4 and Supp. Fig. 5 in the results section of the text (lines 203-238).

(10) Figure S5 also needs a better explanation. E.g., in C-D, what does 'average' mean? The gray is an SD of this average? You modeled a range of values...but which ones are physiological? To me, this is a key point.

We have elaborated on our description of both Supp. Fig. 4 and Supp. Fig. 5 in the results section of the text (lines 203-238).

Reviewer #2 (Recommendations For The Authors):

General:

The images and 3-D reconstructions are visually stunning, but they are not colorblind-friendly and in some cases, hard to distinguish. This shows up particularly in the green and blue colors used in Figure 1. Also, better representative images could be used for Figure 1B.

Thank you for pointing out that blue and green were difficult to distinguish in Figure 1H. We have outlined the green inhibitory puncta in this image to make them more distinguishable. We have also increased the resolution of the image in Figure 1B for better clarity. All other colors are selected from Wong, 2011 (PMID: 21850730; DOI: https://doi.org/10.1038/nmeth.1618).

Supplemental Figure 1D: The low-power view is good to have, but the CN is too small and the image appears a bit noisy. An inset showing the CN on a larger scale (higher resolution image?) would be more convincing. In this image, I see what appear to be cells in the DCN labeled, which calls into question the purity of the source of optogenetic synaptic activation. It is also difficult to tell whether there are other cells labeled in the VCN. Such inputs would still be minor, but it would be good to be very clear about the expression pattern.

To offer more information about the activity of the Ntng1Cre line in other regions of the auditory system, we increased the resolution of the image included in Supp. Fig. 1D and have also included an additional image (Supp. Fig. 1E) of a coronal section of the cochlear nucleus complex with Ntng1-tdT labelling. This image provides additional context for the cells labeled in the DCN. The text in the figure legend has been changed to clarify that some cells in the DCN were labeled (lines 118-120).

We agree that in the Ntng1Cre experiments, there is the possibility of minor contamination from excitatory cells that express ChR2 outside of the spiral ganglion. This is also true for our Foxg1Cre and Foxg1Flp experiments, because these lines label cortical cells in addition to cochlear cells. However, we do not observe direct descending inputs from the cortex into the PVCN, making contamination from other Foxg1Cre-positive neurons unlikely. While non-cochlear inputs from the Ntng1Cre line are possible, evidence from both lines gives us confidence that we are not capturing inputs to octopus cells outside the cochlea. Central axons from Type I spiral ganglion neurons have VGLUT1+ synaptic terminals. When comparing the overlap between VGLUT1+ terminals and Foxg1-tdT labelling, we see full coverage. That is, all VGLUT+ terminals on octopus cells are co-labelled by Foxg1Cre-mediated expression of tdTomato. An example image is shown below. Here, an octopus cell soma is labeled with blue fluorescent Nissl stain and inputs to the cochlear nucleus complex are labeled with Foxg1Cre-dependent tdTomato (Foxg1-tdT; magenta). We have also immunolabeled for VGLUT1 puncta in green. This eliminates the possibility that VGLUT+ cells from outside the cochlea and cortex are sources of excitation to octopus cells.

Further, we have looked at expression of Ntng1-tdT and Foxg1-EYFP together in the octopus cell area. An example image is shown below. All Ntng1-tdT+ fibers (magenta) are also Foxg1-EYFP+ (green), suggesting that all Ntng1Cre-targeted inputs to octopus cells are a part of the Foxg1Cre-targeted input population, which are very likely to only be from the cochlea. We have expanded the results section to include information about the overlap in expression driven by the Ntng1Cre and Foxg1Flp lines.

Supplemental Figure 2 G: These are a bit hard to read. Perhaps use a different image, or provide a reference outline drawing telling us what is what.

We have used a different image with a Thy1-YFP labeled octopus cell for clarity.

In some places, the term "SGN" is used when referencing the axons and terminals within the CN, and without some context, this was occasionally confusing (SGN would seem to refer to the cell bodies). In some places in the text, it may be preferable to separate SGN, auditory nerve fibers (ANFs), and terminals, as entities for clarity.

In order to make the study accessible to a broad neuroscience audience, we refer to the neurons of the spiral ganglion and their central axon projections using one name. We understand why, for those well acquainted with the auditory periphery, condensing terminology may feel awkward. However, for those readers unfamiliar with the anatomy of the cochlea and auditory nerve, we feel that the use of “SGN central axon” makes it clear that the “auditory nerve fibers” come from neurons in the spiral ganglion. This is clarified in the first paragraph of the introduction (lines 29-31) and in the methods (line 533).

Specific: Numbers refer to the line numbers on the manuscript.

L29-31: Cochlear nucleus neurons are more general in their responses than this sentence indicates. While we can all agree that they are specialized to carry (or improve upon) the representation of these specific features of sound, they also respond more generally to sounds that might not have specific information in any of these domains. They are not silos of neural computation, and their outputs become mixed and "re-represented" well before they reach the auditory cortex. Octopus cells are no exception to this. I suggest striking most of the first paragraph, and instead using the first sentence to lead into the second paragraph, and putting the last sentence (of the current first paragraph) at the end of the second (now first) paragraph.

We agree with this assessment and have made major changes to the introduction in line with these suggestions.

L33-46: A number of points in this paragraph need references (exp. line 41).

We agree and have added references accordingly.

L43: Not sure what is meant by "fire at the onset of the sound, breaking it up into its frequency components"?

We changed this text as part of a major reworking of the introduction.

L47-66: Again more citations are needed (at the end of sentence at line 55, probably moving some of the citations from the next sentence up).

We agree and have added references accordingly.

L51: The consistent orientation of octopus cell dendrites across the ANFs has been claimed in the literature (as mentioned here), but there are some (perhaps problematic - plane of sectioning?) counterexamples from the older Golgi-stained images, and even amongst intracellularly stained cells (for example see Reccio-Spinoza and Rhode, 2020). This is important with regards to the broader hypothesis regarding traveling-wave compensation (e.g., McGinley et al; but also many others); if the cells are not all in the appropriate orientation then such compensation may be problematic. Likewise, the data from Lu et al., 2022, points towards a range of sensitivity to frequency-swept stimuli, some of which work in opposition to the traveling wave compensation hypothesis. It would seem that with the Thy1 mice, you have an opportunity to clarify the orientation. Figures 1A and 2A show a consistent dendritic orientation, assuming that these drawings are reconstructions of the cells as they were actually oriented in the tissue. Can you either comment on this or provide clearer evidence?

We are happy to offer more information about the appearances of octopus cells in our preparations. In our hands, sparsely labeled octopus cells in Thy1-YFP-H mice show consistent dendritic orientation when visualized in a 15 degree parasaggital plane, with the most diversity apparent in cells with somas located more dorsally in the octopus cell area. We hypothesize that this is due to the limited area through which the central projections of spiral ganglion neurons (i.e. ANFs) must pass through before they enter the dorsal cochlear nucleus and continue their tonotopic organization in that area.

A caveat to studies without physiological or genetic identification of octopus cells is the assumption that all neurons in the octopus cell area are octopus cells. We find, especially along the borders of the octopus cell area, that stellate cells can be seen amongst octopus cells. Because stellate cell dendrites are not oriented like octopus cell dendrites, any stellate cells misidentified as octopus cells would appear to have poorly-oriented dendrites. This may explain why some studies report this finding. In addition, it can be difficult to assess tonotopic organization because of the 3D trajectory of tightly bundled axons, which is not capturable by a single section plane. Although a parasaggital plane of sectioning captures the tonotopic axis in one part of the octopus cell area, that same plane may be perpendicular at the opposing end.

L67: canonical -> exceptional.

Thank you for the suggestion. We have made this change in the introduction.

L127: This paragraph was confusing on first reading. I don't think Supplemental Figure 1D shows the restricted pattern of expression very clearly. The "restricted to SGNs" might be better as "restricted to auditory nerve fibers" (except in the DCN, where there seem to be some scattered small cells?). A higher magnification image of the CN, but lower magnification than in panel E, would be helpful here.

To avoid confusion, we have re-written this paragraph (lines 117-127) and included a higher magnification image of the CN in a revised Supp. Fig. 1.

L168: Here, perhaps say ANFs instead of SGNs.

As above, we have decided to describe ANFs as SGN central axons to make the anatomy more accessible to people unfamiliar with cochlear anatomy.

L201-204: The IPSPs are surprisingly slow (Figures 5B, C), especially given the speed of the EPSPs/EPSCs in these cells. This is reminiscent of the asymmetry between EPSC and IPSC kinetics in bushy cells (Xie and Manis, 2014). The kinetics used in the model (3 ms; mentioned on line 624) however seem a bit arbitrary and no data is provided for the selection of that value. Were there any direct measurements of the IPSC kinetics (all of the traces in the paper are in the current clamp) that were used to justify this value?

The kinetics of the somatically-recorded IPSPs are subject to the effects of our pharmacological manipulations. EPSPs measured at the soma under control conditions are small amplitude and rapid. With pharmacological reduction of HCN and Kv channels, EPSPs are larger and slower (please see figure in response to a similar question posed by Reviewer #1). We expect that this change also occurs with the IPSP kinetics under pharmacological conditions. Our justification of kinetics has been expanded and justified in the methods section (lines 641-661).

L594: Technically, this is a -11 mV junction potential, but thanks for including the information.

We have corrected this in the text (line 618). Thank you for the close reading of all experimental and methodological details.

L595: The estimated power of the LED illumination at the focal plane should be measured and indicated here.

We measured the power of the LED illumination at the focal plane using a PM100D Compact Power and Energy Meter Console (Thorlabs), a S120C Photodiode Power Sensor (Thorlabs), and a 1000µm diameter Circular Precision Pinhole (Thorlabs). Light intensity at the focal plane ranged between 1.9 and 4.1mW/mm2, corresponding to 6% and 10% intensity on the Colibri5 system. We have reported these measurements in the results section (Lines 621-622).

L609: One concern about the model is that the integration time of 25 microseconds is rather close to the relative shifts in latency. While I doubt it will make a difference (except in the number), it may be worth verifying (spot checks, at least) that running the model with a 5 or 10-microsecond step yields a similar pattern of latency shifts (e.g., Supplementary Figure 5, Figure 5).

Also, it is not clear what temperature the model was executed at (I would presume 35C); this needs to be given, and channel Q10's listed.

We realize that additional information is needed to fully understand the model and have added this to the results and the methods. The synaptic mechanism (.mod) files were obtained from Manis and Campagnola (2018) (PMID: 29331233; DOI: https://doi.org/10.1016/j.heares.2017.12.017). Q10 (3) and temperature (22°C) were also matched to parameters from Manis and Campagnola (2018). Because temperature is a critical factor for channel kinetics, we verified that our primary results remain consistent under conditions using a temperature of 35°C and a time step of 5µs, depicted below. Panel A illustrates the increase in IPSP as a function of glycine conductance under Kv+HCN block conditions at 35°C. As at 22°C, an increase in IPSP magnitude is absent in the control condition at 35°C. Panels B and C provide a direct comparison between the initial (i.e. 22°C) and suggested (i.e. 35°C) simulation conditions. Again we found that temperature does not have a major impact on the amplitude of IPSPs. Thus, results at 35°C do not change the conclusions we make from the model.

The nominal conductance densities should at least be provided in a table (supplemental, in addition to including them in the deposited code). The method for "optimization" of the conductance densities to match the experimental recordings needs to be described; the parameter space can be quite large in a model such as this. The McGinley reference needs a number.

We added a more thorough description of modeling parameters and justification of choices in the methods section of the text (lines 641-661). We have also added a reference number to the McGinley 2012 reference in the text.

I think this is required by the journal:

The model code, test results, and simulation results should be deposited in a public resource (Github would be preferable, but dryad, Zenodo, or Figshare could work), and the URL/doi for the resource provided in the manuscript. This includes the morphology swc/hoc file. The code should be in a form, and with a description, that readily allows an interested party with appropriate skills to download it and run it to generate the figures.

We will upload the code and all associated simulation files to the ModelDB repository upon publication.

Version 2

Abstract

Animals navigate the auditory world by recognizing complex sounds, from the rustle of a predator to the call of a potential mate. This ability depends in part on the octopus cells of the auditory brainstem, which respond to multiple frequencies that change over time, as occurs in natural stimuli. Unlike the average neuron, which integrates inputs over time on the order of tens of milliseconds, octopus cells must detect momentary coincidence of excitatory inputs from the cochlea during an ongoing sound on both the millisecond and submillisecond time scale. Here, we show that octopus cells receive inhibitory inputs on their dendrites that enhance opportunities for coincidence detection in the cell body, thereby allowing for responses both to rapid onsets at the beginning of a sound and to frequency modulations during the sound. This mechanism is crucial for the fundamental process of integrating the synchronized frequencies of natural auditory signals over time.

Introduction

Perception depends on the ability of neurons to encode discrete features of the sensory environment. To generate accurate percepts of complex auditory stimuli, neurons must compute both what frequencies are present and when those frequencies occur across multiple time scales. For instance, overlapping sounds, such as two competing speakers in a noisy room, are distinguished perceptually by correctly binding frequencies with coherent onsets and synchronized temporal modulations1,2. Such computations require coincidence detection that can encode co-occurring frequencies with submillisecond precision. Frequency information is communicated by spiral ganglion neurons (SGNs), whose central axons, also called auditory nerve fibers, project through the eighth nerve, bifurcate, and spread tonotopically to fill each division of the cochlear nucleus complex (CNC)3–7 (Fig. 1A). In addition, SGNs fall into physiologically distinct subtypes that are recruited at different intensities, allowing sounds to be detected across a wide dynamic range 8–14. Many target neurons receive SGN inputs from a limited range of frequencies, encoding only a fraction of the frequency components within a stimulus. This presents a challenge for auditory circuits which must ultimately bind co-occurring frequencies while retaining sequence information in order to locate and recognize sounds. Understanding how these first computations are made is a key step towards deciphering the basis of perception.

Excitatory and inhibitory synapses to octopus cells form two domains.

(A) Illustration of spiral ganglion neuron (SGN) central axons branching within a parasagittal section of the mouse cochlear nucleus complex (CNC). SGN somas in the cochlea are tonotopically organized according to frequency. Axons remain organized throughout the ventral (VCN) and dorsal (DCN) divisions of the CNC. Octopus cells (inset) are found in the octopus cell area (OCA) of the VCN. (B) Excitatory SGN puncta labeled with a VGLUT1 antibody (left) and glycinergic puncta labeled with Glyt2Cre-dependent syp/tdT (Glyt2-syp/tdT; right) in a parasagittal section of the CNC. The teardrop-shaped OCA is not devoid of inhibitory inputs, although they are less prominent than in the surrounding CNC. (C) A Thy1 sparsely labeled octopus cell with excitatory SGN (VGLUT1) and inhibitory (Glyt2-syp/tdT) puncta. Micrographs of 3µm confocal z-stacks show puncta on the medial surface of a soma (top) and a dendrite (bottom). (D) Representative reconstruction of excitatory SGN puncta labeled with Foxg1Cre-dependent syp/tdT (Foxg1-syp/tdT; blue) onto a Thy1 sparsely labeled octopus cell (white). (E) Representative reconstruction of inhibitory puncta labeled with Glyt2-syp/tdT (green) onto a Thy1 sparsely labeled octopus cell (white). (F) Puncta density for excitatory SGN (Foxg1-syp/tdT; black: 10.7 ± 3.0, n = 8 cells, 4 mice) and glycinergic puncta (Glyt2-syp/tdT; green: 4.2 ± 0.8, n = 8 cells, 3 mice) on octopus cell dendrites. Data are presented as mean ± SD. (G) Puncta density on somas for excitatory SGN (black: 13.3 ± 2.2, n = 8 cells, 4 mice) and glycinergic puncta (green: 1.8 ± 0.1, n = 8 cells, 3 mice) and the density along the length of dendrites. Data are presented as mean ± SEM. (H) Top: Illustration of an octopus cell and the ratio between excitatory SGN puncta and glycinergic puncta. Inset: Il-lustration of an octopus cell and the relative innervation densities of excitatory SGNs (blue) and inhibitory puncta (green).

In the mammalian auditory system, precise encoding of broadband timing information begins with the octopus cells of the CNC. Octopus cells are excitatory neurons that bind together co-occurring frequency information on a submillisecond timescale and send this information along one of the parallel ascending pathways in the auditory brainstem. Octopus cells are named for their large-diameter tentacle-like dendrites15,16, which are oriented unidirectionally across a tonotopic array of SGNs such that each neuron integrates inputs from a wide range of frequencies17–20. SGNs provide the major excitatory inputs onto octopus cells. Biophysically, octopus cells have low input resistances near rest (∼4MΩ)21,22, fast time constants (∼200µs)21–23, and large low-voltage-activated potassium (∼40nS at rest)21,23 and hyperpolarization activated (∼62nS at rest)21,22 conductances. Together these properties give octopus cells impressively narrow windows of coincidence detection on the order of 1 millisecond24–28. This combination of receiving SGN innervation from broad frequencies and their biophysical specializations establish octopus cells as spectrotemporal coincidence detectors that can reliably encode the timing of complex stimuli, such as the broadband transients found in speech and other natural sounds21,29,30. Fittingly, in vivo recordings from octopus cells demonstrate their ability to phase lock to broadband transients at rates up to 1kHz31–33. Moreover, computational models of octopus cells demonstrate that onset responses are governed by the cell’s biophysical specializations and are, in large part, the result of temporal summation of excitation34–40. The simplicity of its connectivity combined with the precision of its temporal computations makes the octopus cell an attractive model for understanding how specialized anatomical and electrophysiological properties contribute to neuronal computations.

Although the octopus cell’s integration of SGN inputs within a narrow time frame enables exceptional coincidence detection, such a model does not explain how other temporal features of sound stimuli are encoded. Indeed, octopus cells encode spectrotemporal sequences within their broadly-tuned response areas, like frequency modulated sounds, that likely require further circuit specializations41. Although somatic depolarization can be sufficient to activate an octopus cell35, the vast majority of synapses are found on dendrites. Further, SGN inputs are organized tonotopically along octopus cell dendrites, with inputs from high frequency regions located more distal than those from low frequency regions. Dendritic morphology, passive cable properties, active resting membrane properties, and the spatial relationship between synaptic inputs can all impact excitatory post synaptic potential (EPSP) summation as excitation sweeps across the dendritic arbor and towards the soma. This raises the possibility that computations made in the dendrites influence the effective window of coincidence detection by the octopus cell. Such mechanisms could enable flexible processing that is adaptive to the dynamics of the environment without compromising high-fidelity coincidence computations.

Here, we sought to define the circuit mechanisms that allow octopus cells to act as coincidence detectors across time scales. We generated a comprehensive anatomical and physiological map of excitatory and inhibitory synaptic inputs onto octopus cell somas and dendrites and examined how this circuit organization influences octopus cell activation. Through a combination of in vitro experiments and computational modeling, we show that somatic summation of excitation is shaped by dendritic inhibition. Thus, octopus cells depend on compartmentalized computations that enable preservation of timing information both at the moment of stimulus onset and within an extended window for evidence accumulation, which is fundamental for the spectrotemporal integration of natural auditory stimuli.

Results

The balance of excitatory and inhibitory synapses is different in somatic and dendritic domains

To determine the wiring pattern that drives octopus cell computations, we generated a detailed map of excitatory and inhibitory synaptic inputs (Fig. 1). Overall, octopus cells receive abundant excitatory VGLUT1+ innervation from SGNs42,43 and sparse inhibitory innervation from glycinergic neurons, as visualized using the glycinergic Cre driver Glyt2Creand the Ai34 synaptophysin-tdTomato (syp/tdT) fusion protein reporter (Fig. 1B). The Glyt2- syp/tdT inhibitory inputs nestle between SGN inputs, especially on octopus cell dendrites (Fig. 1C).

Quantification of the number and distribution of presynaptic puncta onto octopus cells revealed marked differences in the ratio of excitation and inhibition in the somatic and dendritic compartments. Since innervation patterns have never been systematically analyzed, we made three-dimensional reconstructions of 16 octopus cells and their excitatory SGN (n = 8 cells, 4 mice) and inhibitory (n = 8 cells, 3 mice) inputs. Octopus cells were visualized using a Thy1 reporter and presynaptic puncta were labeled with the syp/tdT reporter driven either by Foxg1Cre(Fig. 1D) or Glyt2Cre (Fig. 1E). Consistent with qualitative assessment, the density of SGN inputs was higher (10.7 ± 3.0 SGN puncta/100µm2) than that of inhibitory inputs (4.2 ± 0.8 puncta/100µm2, Fig. 1F). Moreover, the relative proportions of excitatory and inhibitory inputs differed in the soma and dendrites (Fig. 1G). On somas, SGNs provided dense innervation that continued on the proximal dendrite, then gradually declined with distance from the soma. By contrast, somas received very few inhibitory inputs. On dendrites, inhibitory puncta were evenly distributed. As a result, octopus cells have a strikingly different average ratio of excitatory and inhibitory puncta on the soma (7:1) and on the dendrite (5:2), suggesting that each compartment contributes differentially to the final computation made by the octopus cell (Fig. 1H).

The majority of excitatory synapses on octopus cells are from type Ia SGNs

Although uniformly glutamatergic, SGNs exhibit stereotyped physiological differences in response thresholds that could affect the nature of their inputs onto octopus cells and hence coincidence detection5,44–46,46,47. There are three molecularly distinct SGN subtypes, referred to as Ia, Ib, and Ic SGNs, which correlate with previously shown physiological groups10,48–53 (Fig. 2A). Therefore, we further categorized excitatory inputs based on SGN subtype identity. These can be identified with the presence of Ntng1Cre-dependent reporter expression54 in Ib and Ic SGNs (Ib/c) and its absence in Ia SGNs50,51,53, coupled with very low to undetectable levels of calretinin (CR-) in Ic SGNs, and moderate to high levels of calretinin (CR+, CR++) in Ib and Ia SGNs (Fig. 2B). As determined using the Ai14 tdTomato (tdT) reporter, Ntng1Cre-labeled Ib/c SGNs accounted for 60.1 ± 2.6% of the entire population, with 28.5 ± 12.2% Ib SGNs, 31.6% Ic SGNs, and 39.9 ± 2.6% Ia SGNs (Fig. 2C: n = 1599 neurons, 4 mice; mean ± SD). These proportions matched scRNA-seq estimates (Fig. 2C), indicating that this approach provides full coverage. SGN subtype identity was further confirmed by examining the spatial organization of SGN peripheral processes en route to the inner hair cells (IHCs) in the cochlea, with Ia processes positioned deeper than Ib and Ic processes (Fig 2A, Supp. Fig. 1A-C).

Type Ia SGNs are the primary excitatory contributors to octopus cells.

(A) Ia (yellow), Ib (orange), and Ic (magenta) spiral ganglion neuron (SGN) axons innervate the cochlear nucleus complex (CNC). SGNs have a continuum of properties organized along the pillar-modiolar axis of inner hair cells (IHCs) and the habenula. Ntng1-expressing Ib/c fibers are positioned on the modiolar side (closest to the ganglion). Strongly calretinin immunopositive Ia fibers are on the other side (closest to the pillar cells). This organization correlates with spontaneous rates (SR) and thresholds (thresh.) measured in vivo. Somas of all SGN subtypes are found at all tonotopic locations. (B) Calretinin (CR) immunolabeling distinguishes SGN subtypes. Ia/b somas label with high (CR++) and medium (CR+) levels of CR, respectively. Ic somas label with low to undetectable levels of CR (CR-). Ntng1Cre-mediated expression of tdT (Ntng1-tdT) labels Ib/c SGNs. (C) Ia SGNs (tdT-, CR++) make up 39.9 ± 2.6% of the SGN population. Ib/c SGNs (tdT+) make up 60.1 ± 2.6% of the SGN population. Ib SGNs (tdT+CR+) make up 28.5 ± 12.2% of the SGN population (n = 1599 neurons, 4 mice). Data are presented as mean ± SD; individual points represent percent coverage per animal, lines connect measurements from the same animal. Dotted lines indicate percentages from 1: Petitpré et al. 2018, 2: Shrestha et al. 2018, and 3: Sun et al. 2018. (D) Representative reconstruction of Ib/c puncta labeled with Ntng1Cre-dependent syp/tdT (Ntng1-syp/tdT; magenta) onto a Thy1 sparsely labeled octopus cell (white). (E) Puncta density for all SGNs (Foxg1-syp/tdT; black: data from Fig. 1F), Ia SGNs (Foxg1 - Ntng1; yellow: 6.6 ± 1.0), and Ib/c SGNs (Ntng1-syp/tdT; magenta: 4.1 ± 1.0, n = 9 cells, 5 mice) along the total dendritic length. Ia density was calculated by subtracting Ib/c density from total SGN density; lines connect measurements from the same reconstruction. Data are presented as mean ± SD. (F) Puncta density on somas for all SGNs (Foxg1-syp/tdT; black: data from Fig. 1E), Ia SGNs (Foxg1 - Ntng1; yellow: 8.4 ± 2.3), and Ib/c SNGs (Ntng1-syp/tdT; magenta: 4.9 ± 1.2, n = 9 cells, 5 mice) and the density along the length of dendrites. Data are presented as mean ± SEM. (G) Illustration of an octopus cell and the relative innervation densities of Ia SGNs (yellow), Ib/c SGNs (magenta), and inhibitory puncta (green).

Within the ventral cochlear nucleus (VCN), where octopus cells reside, Ntng1Cre labeling was restricted to SGN central axons (Supp. Fig. 1D-E). There is sparse labeling in the deep layer of the dorsal cochlear nucleus (DCN) and strong labelling throughout the thalamus, hippocampus, and cortex (Supp. Fig. 1D-E). It is unlikely that the Ntng1Cre labeled cells outside of the cochlea make synapses with octopus cells. When Ntng1Cre driven tdT is co-expressed with Foxg1Flp driven EYFP, all tdT-labeled axons in the VCN also expressed EYFP, suggesting that all our Ntng1Cre labeled inputs are a part of the Foxg1Flp labeled population, which are very likely to only be from SGNs in the cochlea. Additionally, in the octopus cell area, CR+ SGN central axons were segregated from Ntng1-tdT labeled central axons, consistent with the moderate to undetectable levels of CR in Ntng1-tdT SGN somas in the periphery (Supp. Fig. 2F). Thus, Ntng1Cre-driven expression of syp/tdT is an appropriate tool for mapping subtype-specific connectivity onto octopus cells. Reconstruction of Ntng1Cre-labeled Ib/c puncta (Fig. 2D) demonstrated that octopus cells are dominated by inputs from Ia SGN fibers, which have the lowest response thresholds and highest rates of spontaneous activity. Octopus cell dendrites received 4.1 ± 1.0 puncta/100µm2 from Ib/c SGNs (Fig. 2E, magenta: n = 9 cells, 5 mice; mean ± SD), accounting for 38% of the total SGN density. Given that Ntng1-tdT+ cells account for 60.1% of the SGN population (Fig. 2C, magenta), Ib/c inputs were underrepresented on octopus cells. Octopus cells receive similarly low innervation from Ic inputs (Supp. Fig. 2G-I: n = 6 cells, 2 mice), as estimated from the degree of sparse labeling of Ic axons achieved by Myo15iCre-driven reporter expression (Supp. Fig. 2A-F) and the expected proportion of Ic SGNs in the ganglion (Supp. Fig. 2F). By contrast, Ia SGNs, which comprise ∼40% of the total population (Fig. 2C, yellow), accounted for 62 ± 9.7% of SGN synapses on octopus cells (Fig. 2E, yellow: 6.6 ± 1.0 puncta/100µm2). All three subtypes showed the same overall distribution from the soma to the distal dendrite (Fig. 2F). Together, excitatory and inhibitory puncta densities in the innervation maps indicate the average octopus cell receives ∼1035 SGN synapses (642 Ia SGN, 393 Ib/c SGN) and ∼354 inhibitory synapses. Additionally, 83% of the synapses were on the dendrites, suggesting a critical role in the octopus cell computation. Octopus cell reconstructions showed the same basic wiring patterns regardless of where each cell was positioned in the octopus cell area. The tonotopic position of all reconstructed octopus cell somas was estimated in 3D reconstructions aligned to a normalized CNC model of tonotopy55. Octopus cells had similar morphologies (Supp. Fig. 3E-G) and patterns of synaptic innervation (Supp. Fig. 3H-M) regardless of where they were positioned along the tonotopic axis. Thus, Ia SGN synapses are the primary inputs to both the soma and dendrites of octopus cells. Additionally, the whole-neuron wiring diagram identifies a dendritic domain where inhibitory synapses are approximately equal in number to the Ib/c excitatory synapses from the periphery.

SGN inputs to octopus cells facilitate at high stimulation frequencies

Whether or not an octopus cell responds to its inputs depends on when and how EPSPs travel to and then summate in the soma. To determine if SGN subtypes transmit information differently to their central targets, we performed in vitro whole-cell current clamp recordings of octopus cells (Fig. 3A) while using Cre-dependent Channelrhodopsin-2 (ChR2) to evoke EPSPs from all SGNs (Foxg1-ChR2) or only Ib/c SGNs (Ntng1-ChR2). Octopus cells have short action potentials (∼5-15mV) that resemble their large, well-timed EPSPs25,56. Therefore, we evoked small EPSPs that were below spike threshold and distinguishable from action potentials using a phase plot analysis. Trains of ChR2-evoked EPSPs, ranging from 5 to 50Hz, in both the total SGN population (Fig. 3B, black: n = 8 cells, 5 mice) and the Ib/c SGN population (Fig. 3B, magenta: n = 7 cells, 6 mice) exhibited no differences in paired-pulse plasticity at any frequency of stimulation (p > 0.35 at all interstimulus intervals, Tukey’s HSD), although the Ib/c population exhibited higher variability than the total SGN population (at 20ms: SD = 0.11, SD = 0.24, respectively).

SGN subtype inputs to octopus cells do not differ in short term plasticity.

(A) Illustration of the experimental paradigm and representative EPSP traces recorded during in vitro whole-cell current-clamp recordings of octopus cells. Spiral ganglion neuron (SGN) central axon stimulation methods included electrical stimulation or full-field, light-evoked activation of Foxg1- ChR2 or Ntng1-ChR2 SGNs. TTL trigger pulses are shown in gray. (B) Paired pulse ratios for electrically stimulated SGNs (open circles: n = 5 cells, 3 mice), ChR2 stimulated SGNs (Foxg1-ChR2; black: n = 8 cells, 5 mice), and ChR2 stimulated Ib/c SGNs (Ntng1-ChR2; magenta: n = 7 cells, 6 mice) at three interstimulus intervals. With electrical stimulation, SGN inputs to octopus cells were stable and exhibited slight facilitation at 50 Hz (20ms interstimulus interval). ChR2 stimulation caused paired pulse depression not seen with electrical stimulation. Data are presented as mean ± SD. Each data point represents the average paired pulse ratio for a cell. p values from ANOVA and subsequent Tukey HSD test are reported for comparisons between methods of SGN activation (electrical and ChR2) and SGN subpopulation composition within method of activation (SGN-ChR2 and Ib/c-ChR2). Welch’s ANOVA was used for comparisons at 20ms interstimulus interval (50Hz) as data in this condition did not meet the homogeneity of variance assumption. (C) Pulse ratios at 50Hz for electrically stimulated SGNs with physiological 1.4mM Ca2+ ACSF (open circles: n = 5 cells, 3 mice), electrically stimulated SGNs with 2.4mM Ca2+ ACSF (grey: n = 3 cells, 2 mice) and ChR2 stimulated SGNs with physiological ACSF (Foxg1-ChR2; black: n = 8 cells, 5 mice). Data are presented as mean ± SD. p < 0.001 from ANOVA and subsequent Tukey HSD test for all comparisons between methods of SGN activation (electrical and ChR2). There were no statistically significant differences for all comparisons under 1.4mM and 2.4mM Ca2+ (p > 0.100, ANOVA).

ChR2-evoked synaptic responses are known to undergo synaptic depression57,58. To determine if the pairedpulse depression measured in ChR2-stimulated experiments was physiological, we used electrical stimulation to evoke EPSPs from SGNs (Fig. 3B). Electrically-evoked EPSPs had higher paired-pulse ratios than ChR2-evoked EPSPs and were mildly facilitating at short (20ms) intervals (Fig. 3B-C, open circles: n = 5 cells, 3 mice), consistent with an octopus cell’s ability to respond reliably to click trains in vivo17,19,32,59. In contrast, previous results using electrical stimulation demonstrated short-term depression of SGN inputs to octopus cells60,61. However, these experiments were carried out in the presence of higher, non-physiological levels of extracellular calcium. We repeated paired-pulse plasticity experiments with non-physiological calcium concentrations (2.4mM) and similarly found that electrically-evoked EPSPs from SGNs resulted in short-term depression at 50Hz of electrical stimulation (Fig. 3C, grey: n = 3 cells, 2 mice), though not to the degree observed when using full-field, ChR2-evoked inputs.

Glycine evokes inhibitory post synaptic potentials that are occluded by a low input resistance

Given the high density of inhibitory synapses on octopus cell dendrites, we considered the possibility that somatic and dendritic compartments contribute differently to the temporal computation made by octopus cells. A role for inhibition has never been incorporated into octopus cell models as previous efforts failed to reveal physiological evidence of functional inhibitory synapses either in vitro25,56,62 or in vivo41. Similarly, we did not observe lightevoked (Glyt2-ChR2) inhibitory post synaptic potentials (IPSPs) in octopus cell somas during whole-cell current clamp recordings from P30-45 mice. Since inhibitory synapses are located primarily on octopus cell dendrites, we posited that their voltage spread to the soma is limited given the extremely low input resistance of octopus cells.

To decrease electrotonic isolation of the dendrites and increase input resistance, we pharmacologically reduced conductances from voltage-gated potassium (Kv) and hyperpolarization-activated cyclic nucleotide-gated (HCN) channels using 100µM 4-Aminopyridine (4-AP) and 50µM ZD 7288 (ZD). This cocktail increased octopus cell membrane resistance (Fig. 4A) and hyperpolarized the resting membrane potential ∼8-10mV. To compensate for this change, membrane potentials were adjusted to within 3mV of the original resting membrane potential with a holding current. To isolate inhibition, AMPA receptor activation was blocked using 15µM 2,3-dioxo6-nitro-7-sulfamoyl-benzo[f]quinoxaline (NBQX). Consistent with our hypothesis, the increase in input resistance unveiled light-evoked IPSPs in recordings from octopus cell somas (Fig. 4B). Additionally, these IPSPs were fully abolished by bath application of 500nM strychnine (STN) (Fig. 4C), confirming the presence of functional glycinergic inhibitory synaptic transmission onto octopus cells.

Octopus cells receive glycinergic inhibitory post synaptic potentials.

(A) Voltage responses to a -200pA current injection. This representative neuron hyperpolarized 0.7mV (black) in control conditions. After bath application of 100µM 4-Aminopyridine (4-AP) and 50µM ZD 7288 (ZD), hyperpolarizing responses to the same -200pA current injection increased to 8.8mV at steady state (green). (B) Postsynaptic responses to ChR2 stimulation of glycinergic terminals (Glyt2-ChR2) with a 5Hz train (gray) of 1ms full-field blue light pulses before (black) and after bath application of 100µM 4-AP, 50µM ZD, and 15µM NBQX (green: n = 9 cells, 8 mice). Increased input resistance reveals inhibitory potentials that are difficult to detect. (C) Postsynaptic responses to Glyt2-ChR2 stimulation after bath application of 100µM 4-AP, 50µM ZD, and 15µM NBQX (green), with further sequential addition of 20µM picrotoxin (PTX, pink), 100µM cyclothiazide (CTZ, blue), and 500nM strychnine (STN, orange; n = 6 cells, 5 mice). (D) Change in membrane voltage in response to hyperpolarizing somatic current steps in a morphologically and biophysically realistic model of octopus cells before (black) and after removal of voltage-gated potassium (Kv) and hyperpolarization-activated cyclic nucleotide-gated (HCN) channels (blue). As in in vitro current-clamp recordings (A), removing Kv and HCN channels increased the magnitude of voltage responses (ΔVm) to hyperpolarizing current. (E) IPSP magnitude in experimental data (green) and the model (dark and light blue) as a function of input resistance. In somatic measurements, IPSP size increases with input resistance. Modeled IPSPs are shown for two conductance levels (1nS, dark blue; 100nS light blue). (F) IPSPs measured at the soma of a modeled octopus cell before (black) and after removal of Kv and HCN channels (blue). As in in vitro current-clamp recordings, this allows for somatic IPSP detection. (G) Fold change in magnitude of somameasured IPSPs (blue) or dendrite-measured IPSCs at proximal (dark orange) and distal (light orange, open circles) dendritic locations after removal of Kv and HCN channels. Kv and HCN block increases the magnitude of soma-measured IPSPs. The size of dendritic IPSCs are not changed with Kv and HCN block.

To determine the types of glycinergic receptors contributing to IPSPs, we pharmacologically blocked subsets of glycine receptors (Fig. 4C). IPSPs were reduced upon addition of 20µM picrotoxin (PTX), which blocks homomeric glycine receptors63–65. Sequential addition of 100µM cyclothiazide (CTZ), which blocks α2-containing homomeric and heteromeric glycine receptors66,67, nearly abolished the remaining IPSPs, and responses were fully abolished with further application of 500nM STN. These results indicate that relevant glycine receptors include both large conductance extrasynaptic β-subunit lacking homomeric receptors and synaptically localized α2β receptors with slower kinetics68–70.

To confirm whether the electrical confinement of IPSPs to the dendrites is consistent with our understanding of octopus cell biophysics, we developed an improved biophysically and anatomically accurate model of octopus cells based on our findings (Supp. Fig. 4)30,71. Passive and active properties of the model were first aligned with experimental data using a scaling factor (scl), which adjusts the maximum conductance of voltage-gated potassium (Kv) and hyperpolarization-activated cyclic nucleotide-gated (HCN) channels (Supp. Fig. 4A-B). To further align the model with experimental data, we adjusted input resistance (RN) and specific membrane resistance (Rm) to match experimentally measured input resistance under physiological conditions and with Kv and HCN block (Supp. Fig. 4D-E). We further examined the influence of leak, Kv, and HCN conductances on the neuron’s input resistance. In a passive scenario, where leak, Kv, and HCN conductances are set to zero (gleak=0; gKv=0; gh=0), the input resistance scaled proportionally with specific membrane resistance (Supp. Fig F, grey). However, with active Kv and leak conductances, RN was constrained and showed only a slight increase relative to Rm, suggesting attenuation of PSPs during their propagation to the soma (Supp. Fig. 4F-G). As in our current-clamp recordings (Fig. 4A), removal of Kv and HCN conductances in the model changed the input resistance and current-voltage relationship of the neuron, resulting in reduced electrotonic isolation (Supp. Fig. 4C, Fig. 4D, blue). We next tuned the glycine conductance (Ggly) to match experimental results (Fig. 4E). Activation of large (10nS) dendritic glycinergic conductances induced negligible hyperpolarizing voltage changes in the model (Fig. 4F, black). With increased input resistance and reduced electrotonic isolation, dendritic IPSPs measured at the soma were detectable, consistent with our in vitro recordings (Fig. 4F).

While blocking Kv and HCN allowed us to reveal IPSPs at the soma, it is possible that reduced electrotonic isolation does not entirely explain the increase in somaticallymeasured IPSP amplitude. Changes in driving force could increase the magnitude of synaptic currents and therefore increase the magnitude of experimentally measured synaptic potentials. We used the model to explore how changes in membrane potential under Kv and HCN block could impact the magnitude of post synaptic potentials. We simulated dendritic glycinergic conductances over a range of values and in the presence of blocked Kv and HCN channels (Supp. Fig. 4H). Kv and HCN block and the resulting change in input resistance increased the magnitude of somameasured IPSPs for all glycine conductances (Fig. 4G, blue). The size of inhibitory currents measured in the dendrites did not change as a result of the hyperpolarization induced by Kv and HCN block regardless of synaptic location on the dendrites or the magnitude of the glycine conductance (Fig 4G, dark orange, light orange; Supp. Fig 4I-J). These observations suggest that the IPSPs are difficult to detect in physiological conditions because of severe attenuation during transmission along dendrites to the soma. This is further supported by marked changes in transfer impedance (ZT), a measure of signal transmission efficiency, measured with and without Kv and HCN block (Supp. Fig. 4K). Together, results from the model provide evidence of functional glycinergic synaptic transmission that is difficult to detect with in vitro somatic recordings.

Inhibition decreases the magnitude and advances the timing of dendritic SGN inputs

SGN synapses onto octopus cell dendrites are arranged tonotopically, with higher frequency SGNs from the base of the cochlea terminating on distal dendrites and lower frequency SGNs from more apical positions terminating proximally (Fig. 1A). This organization has been proposed to re-synchronize coincidentally firing SGNs that are activated at slightly different times due to the time it takes for the sound stimulus to travel from the base to the apex of the cochlea30. To test how inhibition in the dendrites shapes coincidence detection, we first used our model to explore the influence of simultaneous activation of inhibitory and excitatory synapses at varying locations along the dendritic tree72,73. We modelled how somatically recorded EPSPs are affected by the location of inhibition by moving the site of excitation relative to inhibitory synapses placed either on proximal or distal dendrites (Fig. 5A, Supp. Fig. 5). In our model, inhibitory synapses that are located proximally to excitation (Supp. Fig. 5A) had less of an influence on excitation recorded at the soma compared to those inhibitory synapses located distally to excitation (Supp. Fig. 5B). We analyzed the effect of inhibitory synapse location on somatically measured EPSPs using varying excitatory and inhibitory synaptic weight values (Supp. Fig. 5C-F) and determined that inhibitory conductances (GGly) between 6 and 10nS produced values within experimental ranges. For both inhibition proximal to excitation (Fig. 5B-C) and inhibition distal to excitation (Fig. 5D-E), the model predicted that inhibition reduces EPSP amplitude and accelerates EPSP peak timing at the soma. Thus, the presence of inhibition could modulate EPSP timing in dendritic compartments during continuous auditory stimuli, when inhibition can be recruited after the onset of a sound and thus allow for adaptable temporal processing during the sound’s duration.

Coincident excitation and inhibition on octopus cell dendrites advances EPSP peak times.

(A) The impact of distance between excitatory and inhibitory synapses was measured in a computational model of octopus cells. Inhibitory synapses were placed either proximally or distally to excitation. Excitatory synapses were placed at varying locations along the dendritic arbor to change the anatomical distance (Δd) where d=0 is the location of inhibition (GGly) and d=1 is the condition where excitation (GAMPA) and inhibition are maximally separated. EPSPs were measured at the soma in all conditions (green). (B-E) Quantification of the percent change in soma-measured EPSP magnitude (B, D; GAMPA=5nS) and the shift in EPSP peak timing (C, E; GAMPA=2nS) in models of proximal (B-C) and distal (D-E) inhibition. Inhibitory conductances (GGly) between 6 and 10nS produced values within experimental ranges. Example traces show EPSPs with (green) and without (black) inhibition at d=0 and d=1. Distal dendrites (d=1) have higher local input resistance and lower IPSP attenuation due to the sealed end. Inset scale bars are 1mV, 200ms. (F-H) Coincident stimulation of excitation and inhibition changes EPSP shape. (F) Representative responses to independent stimulation of excitatory spiral ganglion neurons (SGNs; black), independent stimulation of inhibitory inputs (light blue), coincident stimulation of both excitation and inhibition (green), and independent stimulation of excitatory SGNs with the addition of 1µM strychnine (STN). Quantification of the percent change in EPSP height (G) and the shift in EPSP peak timing (H) during coincident Glyt2-ChR2 activation of inhibitory inputs (green: n = 8 cells, 6 mice), bath application of 25µM glycine (dark blue: n = 4 cells, 3 mice), and bath application of 1µM STN (orange: n = 5 cells, 4 mice). Activation of glycinergic receptors during excitation decreased EPSP heights and advanced EPSP peaks. Blocking of tonically active glycine receptors slowed and delayed EPSPs. Data are presented as mean ± SD. Markers represent the average quantification for a cell.

To directly test if the prediction that temporally coincident excitation and inhibition affects the timing and amplitude of excitatory SGN inputs as they travel towards the octopus cell soma, we coincidently activated SGNs and glycinergic inputs in vitro. In these experiments, the octopus cell properties were not altered pharmacologically and inhibition was undetectable or only visible with averaging over many sweeps (Fig. 5F, blue). When synaptic inhibition was evoked together with excitation (Fig. 5F, green), EPSPs recorded in the soma were smaller and faster than when excitation was evoked alone (Fig. 5F, black: n = 8 cells, 6 mice). ChR2-evoked inhibition decreased EPSP heights by 25.2 ± 9.0% (Fig. 5G, green) and shifted the peak of EPSPs forward 57.5 ± 26µs (Fig. 5H, green). This effect was mimicked by bath application of 25µM glycine (Fig. 5G-H, blue: n = 4 cells, 3 mice). Further, bath application of 1µM STN had the opposite effect, resulting in larger EPSPs and delayed peak times (Fig. 5F-H, orange: n = 5 cells, 4 mice). These findings suggest that the timing of EPSP arrival in the soma may be shaped both by tonically active glycine channels and the release of synaptic glycine onto the dendrite. Of note, many SGNs also terminate on the octopus cell soma, where inhibition is minimal. This suggests that the octopus cell’s ability to act as a coincidence detector depends on two stages of compartmentalized computations, one in the dendrite that combines excitation and inhibition to provide important information about which frequencies co-occur in a complex sound stimulus and one in the soma that is restricted by the rigid temporal summation window for coincidence detection. Together with the electrotonic properties of the octopus cell and the dominance of low threshold, low jitter Ia SGN inputs, these combined computations can enable reliable coincidence detection and cross-frequency binding needed for perception of sound.

Discussion

Coincidence detection plays a critical role in many cognitive and perceptual processes, from the ability to localize sound to the binding of auditory and visual features of a common stimulus. Depending on the computation, the temporal window for integration can range widely, thereby requiring circuitry with distinct anatomical and physiological properties. Here, we describe a two-domain mechanism for coincidence detection that can detect co-occurring frequencies with different degrees of precision. By mapping and selectively activating synaptic inputs onto octopus cells both in vitro and in a computational model, we revealed that compartmentalized dendritic nonlinearities impact the temporal integration window under which somatic coincidence detection computations are made. The arrival of many small, reliable excitatory inputs (Fig. 3) rom low-threshold SGNs (Fig. 2) is continuous throughout an ongoing stimulus. We demonstrate that glycinergic inhibition to octopus cell dendrites (Fig. 1) can shift the magnitude and timing of SGN EPSPs as they summate in the soma (Fig. 4,5). The narrow window for coincidence detection computations allows the octopus cell to respond with temporal precision using momentary evidence provided by SGNs at the onset of the stimulus. We propose that, as a stimulus persists, inhibition onto octopus cell dendrites can adjust EPSPs before they arrive at the soma for the final input-output computation. This allows the cell to make an additional computation with a slightly longer window for evidence accumulation without compromising the accuracy of the somatic computation (Fig. 6).

Proposed model of flexible dendritic timing for precise somatic coincidence detection.

(A-B) Spectrograms show how co-occurring frequencies in broadband onsets (shown for a clicking sound, A) and frequency modulations (shown for human squeaking, B), change in their strength (top) and total intensity (bottom) on different time scales. Despite the timing differences, both kinds of stimuli are hypothesized to result in somatic summation during the octopus cell’s spike integration window. (C) Excitation must summate at the soma of octopus cells within a narrow time window (∼1ms) to achieve a depolarization rate rapid enough to trigger action potentials. Summation of excitation alone accounts for responses to sound onsets (inset, top). During stimuli that require summation over a longer time period, such as frequency modulated sweeps, synaptic inhibition can accelerate EPSPs as they travel along octopus cell dendrites towards the soma for coincidence detection computations (inset, bottom). We propose a mechanism for preferential processing of a subset of excitatory inputs where selective temporal advancement of a subset of EPSPs by local inhibition could expand the effective window for coincidence detection at the soma.

As coincidence detectors in the auditory system, octopus cells are faced with the challenge of recognizing complex sounds that include many frequencies that co-occur from the beginning to the end of the stimulus. As shown Flexible Coincidence Detection in the Auditory System by in vivo recordings18,41, octopus cells respond well to cues that include complex spectrotemporal patterns, including frequency modulations beyond the onset of the stimulus41. Given that the auditory environment is filled with overlapping sound stimuli, such responses presumably allow the octopus cell to encode which frequencies belong to which sound. Our data thus support the role of octopus cells beyond simple onset coincidence detectors that rely solely on the temporal summation of excitation. The results suggest that, in addition to high Kv and HCN conductances at rest, the addition of dendritic inhibition transforms the magnitude and timing of SGN signals as they arrive in the cell body, which may expand the response selectivity of an octopus cell and allow them to become a slightly leakier integrator that can accumulate evidence beyond onsets. Although, this inhibition is difficult to detect because of shunting, our data demonstrate that it is both present and impactful.

As well as needing to work beyond onsets, an effective coincidence detector in the auditory system must also function reliably across a range of sound intensities. Intensity information is encoded by the number and types of SGNs that are activated in the cochlea. The Ia, Ib, and Ic molecular subtypes defined in mouse50,51,53 broadly correspond to the anatomically and physiologically defined subtypes described across species52,74. We find that the majority of inputs onto octopus cells come from Ia SGNs, which most closely correspond to the low-threshold, highspontaneous rate population. Consistent with this result, single-unit SGN recordings in cats demonstrated a bias towards low-threshold, high-spontaneous rate axon collaterals in the octopus cell area44. Low-threshold SGNs are also characterized by short first spike latencies and low temporal jitter75–78. A hallmark of the octopus cell is the fact that it only fires action potentials when many SGN inputs are activated within a narrow period of time79. The presence of many low-threshold and temporally precise inputs on the octopus cell may help ensure that coincidence detection still works reliably for quiet sounds. Further, Ia inputs onto octopus cells do not exhibit pairedpulse depression, similar to low levels of depression seen in Ia inputs to bushy cells80. The presence of SGN inputs without paired-pulse depression could be beneficial for encoding sustained auditory signals. Finally, while Ia SGNs are over-represented, Ib and Ic inputs are also present. Since precise, low-threshold SGN responses can be saturated by background noise such that responses to relevant stimuli are masked10,75,78,81, recruitment of higher threshold SGNs at higher sound intensities may compensate for this tradeoff.

The presence of inhibitory inputs onto dendrites is a fundamental feature of the nervous system and, in other systems, contributes to a neuron’s final computation. For example, direction selectivity computations in dendrites of retinal cells require excitation-inhibition interactions in dendritic compartments82,83. In pyramidal cells of the cortex and hippocampus, the spatial distribution of inhibition impacts dendritic non-linearities in a branch-selective manner73,84–88. However, octopus cells do not share all mechanisms for dendritic computation seen in cortical and hippocampal neurons. Resting conductances and low-threshold potassium conductances may suppress voltage-gated calcium spikes and attenuate the magnitude of action potentials as they backpropagate through the soma and dendritic tree21,89–91. Therefore, subthreshold dendritic integration of coincident excitation and local inhibition may be the primary computation that occurs in octopus cell dendrites before action potential generation near the soma.

Although this work uncovers a role for inhibition, a deeper understanding of octopus cell computations will require determining what information is carried by inhibitory inputs. Despite the well-established presence of presynaptic glycinergic puncta in the octopus cell area92–95 and glycinergic receptor expression in octopus cells96–100, it is unknown where glycinergic innervation originates. Local neurons within the CNC could provide inhibition; however it remains unclear whether octopus cells receive connections from D-Stellate56, L-Stellate101, or tuberculoventral cells102. Outside of the CNC, terminal degeneration experiments in cats suggested the superior periolivary nucleus (SPON) and the ventral division of the lateral lemniscus (VNLL) as potential sources of descending inhibition to the octopus cell area103. Octopus cells provide excitatory input to both the SPON104–108 and the VNLL32,109–113, raising the possibility of feedback inhibition from the auditory brainstem as a circuit mechanism that elongates temporal summation windows during ongoing stimuli. Such descending feedback inhibition is not rapid enough to prevent or alter the characteristic octopus cell onset response, but could change the effective coincidence detection window as the stimulus continues or limit the duration of the response. Future studies will be required to identify the source of inhibition and its organization within dendritic compartments. If inhibitory inputs tonotopically match the local, narrowly-tuned dendritic SGN inputs, it is possible that frequency-matched inhibition could influence spectral selectivity or feature extraction. On the other hand, broadly tuned inhibition could reduce depolarization block or serve as a temporal milestone that signals gaps or offsets. Further characterization of in vivo octopus cell responses in complex sound environments may clarify the effect of noise on signal detection and reveal additional features of this cell’s contributions to perception of the auditory world.